---
title: "Combined_RNAseq_Proteomics"
format: html
editor: visual
---

# Loading Libraries

```{r}
library(jsonlite)
library(httr) #For REST API calls
library(tidyverse)
library(ggplot2) #Plotting
library(DESeq2) #DE Analysis
library(Biobase) 
library(gplots)
library(RColorBrewer) 
library(pheatmap) #For heat maps and quality control graphs
library(matrixStats)
library(lattice)
library(gtools)
library(dplyr)
library(edgeR) #For data analysis and limma
library(BiocParallel) #For running DESeq2 in parallel
library(ggsignif)
library(ggpubr) #Adding significance bars for graphs
```

# Loading Data

## TCGA RNA-seq

Done by downloading tsv files from TCGA database and then formatting them by removing the folders each file comes in using these WIndows commands:

To move all files in subfolders into the mainfolder: (while in the parent directory)

for /r %i in (\*) do \@move "%i" .

To remove empty directories:

for /d %i in (\*) do @rmdir /s /q "%i"

### Lung

```{r}
#Loads tsv files of lung cells and appends vertically
#Normal Data
files1L <- list.files("data/Normal_Lung", pattern ="*.tsv", full.names=TRUE) #Has path to folder containing files
col_name = c("gene_id","gene_name","gene_type","unstranded","stranded_first","stranded_second","tpm_unstranded","fpkm_unstranded","fpkm_uq_unstranded")
lung_normal <- read_tsv(files1L, comment="#", col_names = col_name, skip=6)

#Cancer Data
files2L <- c(
  list.files("data/Cancer_Lung/Adenomas", pattern="*.tsv", full.names=TRUE), #Has path to folders containing files
  list.files("data/Cancer_Lung/Squamous", pattern="*.tsv", full.names=TRUE)           
)             

lung_cancer <- read_tsv(files2L, comment="#", col_names = col_name, skip=6)

#Load metadata and select useful columns
metadata <- fromJSON("data/Cancer_Lung/metadata.repository.lung.json")

temp <- metadata |> select(associated_entities, file_name)
temp$case <- bind_rows(temp$associated_entities)$case_id
temp$sample <-sapply(temp$file_name, function(data) str_split_i(data, "\\.", 1))
meta <- temp |> select(sample, case)


#Find list of samples
normList <- sapply(files1, function(data) str_split_i(data, "/", 3))
normList <- sapply(normList, function(data) str_split_i(data, "\\.", 1))

cancList <- sapply(files2, function(data) str_split_i(data, "/", 4))
cancList <- sapply(cancList, function(data) str_split_i(data, "\\.", 1))

sampleList <- c(normList, cancList)

#Get metadata only within the list of samples
lung_metadata <- meta[meta$sample %in% sampleList, ]
```

### Breast

Used if the expression data is in TCGA tsv files and not formatted into rds files

```{r}
# #Load the metadata
# filePath = "data/Normal_Breast/TCGA_BRCA/metadata.repository.2025-06-16.json"
# 
# temp <- fromJSON(filePath)
# 
# #Select columns of interest
# temp <- temp |> select(associated_entities, file_name)
# 
# #Extract barcode from associated entities
# temp$associated_entities <- bind_rows( #Combines all data frames in the list
#                                       temp$associated_entities
#                             )$entity_submitter_id
# 
# #Gets the plate value, which is the 6th part of the barcode
# temp <- temp |> mutate(
#   plate = sapply(
#     temp$associated_entities,
#     function(data) str_split_i(data, "-", 6)
#   )
# )
# 
# #Remove the barcode column and save the file for easier loading
# temp2 <- temp |> select(!associated_entities)
# write_rds(temp2, "data/Normal_Breast/TCGA_BRCA/metadata_BRCA_normal.rds")


# #Reads all the tsv files in a specific order and joins them vertically
# #temp2 was metadata from before
# tempFiles <- sapply(#sapply applies the function over the list
#   temp2$file_name,
#   function(data) paste0("data/Normal_Breast/TCGA_BRCA/", data) #Custom function bc can onyl have one input
# )
# 
# #Use already known col names from TCGA tsv files
# col_name = c("gene_id","gene_name","gene_type","unstranded","stranded_first","stranded_second","tpm_unstranded","fpkm_unstranded","fpkm_uq_unstranded")
# 
# #Reads in data.. Skips the first 6 lines bc of non-data
# tempData <- read_tsv(tempFiles, col_names = col_name, skip = 6)
# 
# #Remove unnecessry columns (Can save tpm data if not using DESeq2)
# tempData <- tempData |> select(gene_id, gene_name, unstranded)
# 
# #Labels each sample using column of sample name
# tempSample <- sapply(
#   temp2$file_name,
#   function(data) str_split_i(data, "\\.", 1)
# )
# 
# #Labels each sample via a number and puts in the file name
# tempData <- tempData |> mutate(
#   sample = (row_number()-1) %/% 60660 + 1
# )
# 
# tempData$sample <- tempSample[tempData$sample]

# #Turns the data into a matrix
# tempMatrix <- tempData |> pivot_wider(names_from = sample, values_from = unstranded)
# 
# 
# #Removes the values after the dots from the gene_ids while keeping the _PAR_Y data
# #Splits the gene_id along the periods 
# temp <- sapply(tempMatrix$gene_id, function(data) str_split_1(data, "\\.")) 
# 
# #Finds the values whose values after the period are longer than 2, since those have _PAR_Y after 
# temp2 <- sapply(temp[2, ], str_length) > 2  
# 
# #Changes the first column to include the _PAR_Y for those values 
# temp[1, temp2] <- sapply(temp[1, temp2], function(data) paste0(data, "_PAR_Y"))  
# 
# #Puts it into the cancer data 
# tempMatrix$gene_id <- temp[1, ]
# 
# #Saves the data in an R RDS file for easier loading
# write_rds(tempMatrix, "data/Normal_Breast/TCGA_BRCA/normal_breast_BRCA.rds")
```

Loading RDS files

```{r}
#TCGA BRCA data
breast_normal <- read_rds("data/Normal_Breast/TCGA_BRCA/normal_breast_BRCA.rds")

#Loads metadata such as batch data for these samples
breast_metadata_normal <- read_rds("data/Normal_Breast/TCGA_BRCA/metadata_BRCA_normal.rds")

#Loads the metadata
breast_metadata_cancer = read_rds("data/Cancer_Breast/metadata_BRCA_cancer.rds")

#Load mRNA sequencing data
breast_cancer <- read_rds("data/Cancer_Breast/cancer_breast_BRCA.rds")
```

## ProteomicsDB

### Lung

### Breast
