---
title: "Limma vs DESeq2"
format: html
editor: visual
---

# Loading libraries

```{r}
#| include: false

library(jsonlite)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library("Biobase")
library("gplots")
library("ggplot2")
library("RColorBrewer")
library("pheatmap")
library("matrixStats")
library("lattice")
library("gtools")
library("dplyr")
library(edgeR)
library(ggsignif)
library("BiocParallel")
library("ggpubr")

```

# Control Variables

```{r}
nN = 20
nC = 20
geneList = c("ADGRE5", "CD109", "CD151", "ALCAM", "CD276", "EPCAM")
all = T
paired = T
runDESeq = F
```

# Functions

```{r}
#To get gene name, run:
#(cancer_gene_key |> filter(gene_name=="CD109"))[1, ]$gene_id
#Plots a boxplot of the gene of interest
geneBoxPlot <- function(geneName, ddsData, resData) {
  geneId <- (cancer_gene_key |> filter(gene_name==geneName))[1, ]$gene_id
  cPlot <- plotCounts(ddsData, 
                    gene=geneId, 
                    intgroup="condition", 
                    returnData = TRUE,
                    transform = TRUE
                    )
  #Gets the adjusted p values for the gene
  pGene <- resData[geneId, ]$padj

    graph <- cPlot |> ggplot(aes(x=condition, y=count, fill=condition))+
    geom_boxplot(outliers = FALSE) + 
#  geom_point(position=position_jitter(w=0.1,h=0), alpha = 0.2) +
  labs(
    title=paste("mRNA Counts for",geneName),
    subtitle=paste("Gene ID:", geneId),
    caption=paste("p-value =", pGene)
  )
  return(cPlot)
}
```

```{r}
#Return points to plot for normal vs cancer
genePlotPoints <- function(geneName, ddsData, resData) {
  geneId <- (cancer_gene_key |> filter(gene_name==geneName))[1, ]$gene_id
  cPlot <- plotCounts(ddsData, 
                    gene=geneId, 
                    intgroup="condition", 
                    returnData = TRUE
                    )
  return(cPlot)
}
```

```{r}
#Return log2cpm points for DESeq2
genePlotPointsLCPM <- function(geneName, ddsData, resData) {
  geneId <- (cancer_gene_key |> filter(gene_name==geneName))[1, ]$gene_id
  data_mat <- assay(normTransform(ddsData))
  gene_data <- data_mat[geneId, ]
  
  pointsData <- as.data.frame(gene_data)

  pointsData$condition <- ddsData$condition
  
  return(pointsData)
}
```

```{r}
#Converts a gene name to an id
nameToID <- function(geneName){
  geneId <- (cancer_gene_key |> filter(gene_name==geneName))[1, ]$gene_id
  geneId <- as.vector(geneId)[1]
  return(geneId)
}
```

# Limma vs DESeq2

This is setting up the combined data and metadata as well as random subsets \### Data loading

```{r}
#Normal Data loading
#TCGA BRCA data
normal_breast <- read_rds("data/Normal_Breast/TCGA_BRCA/normal_breast_BRCA.rds")

#Loads metadata such as batch data for these samples
metadata_normal <- fromJSON("data/Normal_Breast/TCGA_BRCA/metadata.repository.BRCA.normal.json")


#Cancer Data
#Loads the metadata
metadata_cancer = fromJSON("data/Cancer_Breast/metadata.repository.BRCA.cancer.json")

#Load mRNA sequencing data
cancer_breast <- read_rds("data/Cancer_Breast/cancer_breast_BRCA.rds")
```

Modify metadata

```{r}
#Select columns of interest
metTempN <- metadata_normal |> select(associated_entities, file_name)

#Extract barcode from associated entities
metTempN$barcode <- bind_rows( #Combines all data frames in the list
                                      metTempN$associated_entities
                            )$entity_submitter_id

#Extract case id from associated entities
metTempN$subject <- bind_rows( #Combines all data frames in the list
                                      metTempN$associated_entities
                            )$case_id

#Select columns of interest
metTempC <- metadata_cancer |> select(associated_entities, file_name)

#Extract barcode from associated entities
metTempC$barcode <- bind_rows( #Combines all data frames in the list
                                      metTempC$associated_entities
                            )$entity_submitter_id

#Extract cacse id from associated entities
metTempC$subject <- bind_rows( #Combines all data frames in the list
                                      metTempC$associated_entities
                            )$case_id

#Gets the plate value, which is the 6th part of the barcode
metTempN <- metTempN |> mutate(
  plate = sapply(
    metTempN$barcode,
    function(data) str_split_i(data, "-", 6)
  )
)
metTempC <- metTempC |> mutate(
  plate = sapply(
    metTempC$barcode,
    function(data) str_split_i(data, "-", 6)
  )
)


#Finalize metadata
meta_normal <- metTempN |> select(!associated_entities)
meta_cancer <- metTempC |> select(!associated_entities)
```

### Combining Data

Now the data is going to be combined

```{r}
#TCGA
norm_gene_key <- normal_breast |> select(gene_id, gene_name)
tempN <- normal_breast |> select(!c(gene_name))

#Now for cancer samples
cancer_gene_key <- cancer_breast |> select(gene_id, gene_name)
tempC <- cancer_breast |> select(!c(gene_name))

#Name and gene_id are kept to be used as the key for joining them together
#Combines gene id for the combined matrix
combined_breast <- tempN |> full_join(tempC, join_by(gene_id == gene_id))

#Forms the matrix while movign the Name column into the row names
combined_breast_matrix <- as.matrix(
  column_to_rownames(combined_breast, var = "gene_id") #would be "Name" if GTex
)

#Turns NA values to 0
combined_breast_matrix[is.na(combined_breast_matrix)] <- 0


```

Now metadata for all the samples is going to be created

```{r}
#Labels each metadata by the condition
metTempN <- meta_normal |> mutate(condition = "normal")
metTempC <- meta_cancer |> mutate(condition = "cancer")

combined_meta <- bind_rows(metTempN,metTempC)

#Turn file name into sample
combined_meta$sample <- sapply(
  combined_meta$file_name,
  function(data) str_split_i(data, "\\.", 1)
)

combined_meta <- combined_meta |> select(sample, condition, subject, plate)

#Gets subtypes
combined_subtype <- read_rds("data/Cancer_Breast/BRCA_subtype_manual.rds")
temp <- rownames_to_column(combined_subtype) |> select(rowname, subtype = PAM50)
combined_meta <- full_join(combined_meta, temp, join_by(sample == rowname))

#Factors all columns and orders them
combined_meta$sample <- factor(combined_meta$sample)
combined_meta$condition <- factor(combined_meta$condition, levels = c("normal", "cancer"))
combined_meta$subject <- factor(combined_meta$subject)
combined_meta$plate <- factor(combined_meta$plate)
combined_meta$subtype <- factor(combined_meta$subtype)

```

### Random Sampling

```{r}
#Takes a random sample of n points for normal and cancer samples
#50-50
#IF want to use all samples

if (all){
  sub_matrix = combined_breast_matrix
  sub_meta = combined_meta
} else {
  if(!paired){
    #Gets nN random normal samples
    indexs1 <- sample(1:dim(meta_normal)[1], size = nN, replace = FALSE)
    #Gets nC random cancer samples
    indexs2 <- sample((min=dim(meta_normal)[1] + 1):dim(combined_meta)[1], size = nC, replace = FALSE)
    
    #Combined indixes
    indexs <- sort(c(indexs1, indexs2))
    
    #Creates subarrays of the selected samples
    sub_matrix <- combined_breast_matrix[, indexs]
    sub_meta <- combined_meta[indexs, ]
  } else {
  
    #Tries to get nN amount of cases and pairs up normal and cancer samples
    cases <- combined_meta |> filter(condition == "normal") |> select(subject)
    
    indexsP <- sample(1:dim(cases)[1], size = nN, replace = FALSE)
    casesP <- unfactor(cases[indexsP, ])
    
    sub_meta <- combined_meta |> filter(subject %in% casesP) |> group_by(condition, subject) |> slice_head(n = 1) |> ungroup()
  }
    sub_matrix <- combined_breast_matrix[, sub_meta$sample]
}

#Change levels
sub_meta$sample <- factor(unfactor(sub_meta$sample))
sub_meta$subject <- factor(unfactor(sub_meta$subject))
sub_meta$plate <- factor(unfactor(sub_meta$plate))
sub_meta$subtype <- factor(unfactor(sub_meta$subtype))
```

## BC Unpaired

Compared DESeq2 and Limma

### DESeq2 Analysis

```{r}
if (runDESeq){
  
  #Starts counting time
  time1 = Sys.time()
  
  #Uses DESeq2 method to analyze the matrix data
  dds <- DESeqDataSetFromMatrix(countData = sub_matrix,
                                colData = sub_meta,
                                #design = ~ batch + condition)
                                design = ~ condition)
  
  #Sets normal patients as the control
  dds$condition <- relevel(dds$condition, ref = "normal")
  
  #Prefilters to speed up method
  temp <- sub_meta |> group_by(condition) |> summarize(n=n())
  
  smallestGroupSize <- 3
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  #Used reduced method
  dds <- DESeq(dds, fitType = "glmGamPoi", test="LRT", reduced = ~1)
  
  res <- results(dds)
  
  #Now the results of the genes are plotted
  
  #Gathers all plotting points for all genes
  combBC <- data.frame()
  for (gene in geneList){
    combBC <- combBC |> bind_rows(
    genePlotPoints(gene, dds, res) |> mutate(gene=gene))
  }
  
  combBC <- combBC |> mutate(log2count = log2(count))
  
  #plots the values
  graph1D <- combBC |> ggplot(aes(x=gene, y=log2count, fill=condition))+
      geom_boxplot(outliers = FALSE) + 
    labs(
      title=paste("mRNA Counts for genes of interest"),
      subtitle = "DESeq2 Unpaired"
    )
  
  
  
  #Ends the time and prints time elapsed
  
  time2 = Sys.time()
  
  #Prints time elapsed
  tDif1D = time2-time1
}
```

### Limma Analysis

```{r}
#Start time
time1 = Sys.time()

factor1 <- sub_meta$condition

#Makes DGEList object
d0 <- DGEList(sub_matrix)

#Calculates normalization factors
d0 <- calcNormFactors(d0)

#Filter low expression genes
cutoff <- 3
drop <- which(apply(cpm(d0), 1, max) < cutoff)
d <- d0[-drop,] 

#gets lcpm data
logcpm <- cpm(d, prior.count=2, log=TRUE)

#Makes design of analysis
mm <- model.matrix(~0 + factor1)

#VoomLmFit
fit <- voomLmFit(d, mm)
fit <- eBayes(fit)

cPlot = data.frame()

for( gene in geneList){
  norm <- as.data.frame(fit$EList$E[nameToID(gene), ])
  norm$condition <- sub_meta$condition
  norm$gene <- gene
  
  colnames(norm) <- c("counts", "condition", "gene")
  
  cPlot <- cPlot |> bind_rows(norm)
}

# #Plotting expression values
# norm <- as.data.frame(edgeR::cpm(d)[gene, ]) # to get normalized expression values.
# norm$condition <- sub_meta$condition
# colnames(norm) <- c("counts", "condition")

graph1L <- cPlot |> ggplot(aes(x = gene, y= counts, fill = condition)) + geom_boxplot(outliers = F) + 
  labs(
    title=paste("mRNA Counts for genes of interest"),
    subtitle = "Limma-voom Unpaired",
    y = "log2CPM",
  )

#ENd time
time2 = Sys.time()

tDif1L <- time2-time1
```

### Comparison

```{r}
if (runDESeq){
  graph1D
  tDif1D
}

graph1L
tDif1L
```

## BC Plate

Compared DESeq2 and Limma

### DESeq2 Analysis

```{r}
if (runDESeq){
  
  #Starts counting time
  time1 = Sys.time()
  
  #Uses DESeq2 method to analyze the matrix data
  dds <- DESeqDataSetFromMatrix(countData = sub_matrix,
                                colData = sub_meta,
                                #design = ~ batch + condition)
                                design = ~ plate + condition)
  
  #Sets normal patients as the control
  dds$condition <- relevel(dds$condition, ref = "normal")
  
  #Prefilters to speed up method
  temp <- sub_meta |> group_by(condition) |> summarize(n=n())
  
  smallestGroupSize <- 3
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  #Used reduced method
  dds <- DESeq(dds, fitType = "glmGamPoi", test="LRT", reduced = ~plate)
  
  res <- results(dds)
  
  #Now the results of the genes are plotted
  
  #Gathers all plotting points for all genes
  combBC <- data.frame()
  for (gene in geneList){
    combBC <- combBC |> bind_rows(
    genePlotPoints(gene, dds, res) |> mutate(gene=gene))
  }
  
  combBC <- combBC |> mutate(log2count = log2(count))
  
  #plots the values
  graph2D <- combBC |> ggplot(aes(x=gene, y=log2count, fill=condition))+
      geom_boxplot(outliers = FALSE) + 
    labs(
      title=paste("mRNA Counts for genes of interest"),
      subtitle = "DESeq2 Batch"
    )
  
  
  
  #Ends the time and prints time elapsed
  
  time2 = Sys.time()
  
  #Prints time elapsed
  tDif2D <- time2-time1
}
```

### Limma Analysis

```{r}
#Start time
time1 = Sys.time()

factor1 <- sub_meta$condition
plate <- sub_meta$plate


#Makes DGEList object
d0 <- DGEList(sub_matrix)

#Calculates normalization factors
d0 <- calcNormFactors(d0)

#Filter low expression genes
cutoff <- 3
drop <- which(apply(cpm(d0), 1, max) < cutoff)
d <- d0[-drop,] 

#gets lcpm data
logcpm <- cpm(d, prior.count=2, log=TRUE)

#Makes design of analysis
mm <- model.matrix(~plate + factor1 )

#VoomLmFit
fit <- voomLmFit(d, mm)
fit <- eBayes(fit)

cPlot = data.frame()

for( gene in geneList){
  norm <- as.data.frame(fit$EList$E[nameToID(gene), ])
  norm$condition <- sub_meta$condition
  norm$gene <- gene
  
  colnames(norm) <- c("counts", "condition", "gene")
  
  cPlot <- cPlot |> bind_rows(norm)
}

# #Plotting expression values
# norm <- as.data.frame(edgeR::cpm(d)[gene, ]) # to get normalized expression values.
# norm$condition <- sub_meta$condition
# colnames(norm) <- c("counts", "condition")

graph2L <- cPlot |> ggplot(aes(x = gene, y= counts, fill = condition)) + geom_boxplot(outliers = F) + 
  labs(
    title=paste("mRNA Counts for genes of interest"),
    subtitle = "Limma-voom Batch",
    y = "log2CPM",
  )

#ENd time
time2 = Sys.time()

tDif2L <- time2-time1


#Removing batch effects from the plate
#Getting list of samples
cPlot$sample <- c(rep(combined_meta$sample, length(geneList)))

#Makes matrix of log2 genes exp values
tempCounts <- cPlot |> select(!condition) |> pivot_wider(names_from = sample, values_from = counts)
tempCounts <- column_to_rownames(tempCounts, "gene")

#Makes the factor a numeric matrix and removes batch effects
levels(factor1) <- 1:length(levels(factor1))
normCounts <- removeBatchEffect(as.matrix(tempCounts), batch = plate, design = as.matrix(as.numeric(unfactor(factor1))))

#combines all the gene values for all genes into a single list
temp <- c()
for(i in 1:dim(normCounts)[1]){
  temp <- c(temp, normCounts[i, ])
}
temp <- as.vector(temp)

#
cPlot$counts <- temp

graph21L <- cPlot |> ggplot(aes(x = gene, y= counts, fill = condition)) + geom_boxplot(outliers = F) + 
  labs(
    title=paste("mRNA Counts for genes of interest"),
    subtitle = "Limma-voom Batch",
    y = "log2CPM",
  )
```

### Comparison

```{r}
if (runDESeq){
  graph2D
  tDif2D
}

graph2L
graph21L
tDif2L
```

## BC Patient

Compared DESeq2 and Limma

```{r}
if(!paired){
  print("ERROR, paired is FALSE, make TRUE for no error")
}
```

### DESeq2 Analysis

```{r}
if(!paired){
  
  if (runDESeq){
    
    #Starts counting time
    time1 = Sys.time()
    
    #Uses DESeq2 method to analyze the matrix data
    dds <- DESeqDataSetFromMatrix(countData = sub_matrix,
                                  colData = sub_meta,
                                  design = ~ subject + condition)
    
    #Sets normal patients as the control
    dds$condition <- relevel(dds$condition, ref = "normal")
    
    #Prefilters to speed up method
    temp <- sub_meta |> group_by(condition) |> summarize(n=n())
    
    smallestGroupSize <- 3
    keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
    dds <- dds[keep,]
    
    #Used reduced method
    dds <- DESeq(dds, fitType = "glmGamPoi", test="LRT", reduced = ~subject)
    
    res <- results(dds)
    
    #Now the results of the genes are plotted
    
    #Gathers all plotting points for all genes
    combBC <- data.frame()
    for (gene in geneList){
      combBC <- combBC |> bind_rows(
      genePlotPoints(gene, dds, res) |> mutate(gene=gene))
    }
    
    combBC <- combBC |> mutate(log2count = log2(count))
    
    #plots the values
    graph3D <- combBC |> ggplot(aes(x=gene, y=log2count, fill=condition))+
        geom_boxplot(outliers = FALSE) + 
      labs(
        title=paste("mRNA Counts for genes of interest"),
        subtitle = "DESeq2 Patient"
      )
    
    
    
    #Ends the time and prints time elapsed
    
    time2 = Sys.time()
    
    #Prints time elapsed
    tDif3D <- time2-time1
  }
}
```

### Limma Analysis

```{r}
if(!paired){
  
  #Start time
  time1 = Sys.time()
  
  factor1 <- sub_meta$condition
  plate <- sub_meta$plate
  subject <- sub_meta$subject
  
  
  #Makes DGEList object
  d0 <- DGEList(sub_matrix)
  
  #Calculates normalization factors
  d0 <- calcNormFactors(d0)
  
  #Filter low expression genes
  cutoff <- 3
  drop <- which(apply(cpm(d0), 1, max) < cutoff)
  d <- d0[-drop,] 
  
  #gets lcpm data
  logcpm <- cpm(d, prior.count=2, log=TRUE)
  
  #Makes design of analysis
  mm <- model.matrix(~0 + factor1 + subject)
  
  #VoomLmFit
  fit <- voomLmFit(d, mm)
  fit <- eBayes(fit)
  
  cPlot = data.frame()
  
  for( gene in geneList){
    norm <- as.data.frame(fit$EList$E[nameToID(gene), ])
    norm$condition <- sub_meta$condition
    norm$gene <- gene
    
    colnames(norm) <- c("counts", "condition", "gene")
    
    cPlot <- cPlot |> bind_rows(norm)
  }
  
  # #Plotting expression values
  # norm <- as.data.frame(edgeR::cpm(d)[gene, ]) # to get normalized expression values.
  # norm$condition <- sub_meta$condition
  # colnames(norm) <- c("counts", "condition")
  
  graph3L <- cPlot |> ggplot(aes(x = gene, y= counts, fill = condition)) + geom_boxplot(outliers = F) + 
    labs(
      title=paste("mRNA Counts for genes of interest"),
      subtitle = "Limma-voom Patient",
      y = "log2CPM",
    )
  
  #ENd time
  time2 = Sys.time()
  
  tDif3L <- time2-time1
}
```

### Comparison

```{r}

if (runDESeq){
  graph3D
  tDif3D
}

graph3L
tDif3L
```

## BC Subtype + Patient

Compared DESeq2 and Limma

```{r}
if(!paired){
  print("ERROR, paired is FALSE, make TRUE for no error")
}
```

### DESeq2 Analysis

```{r}
if (runDESeq){
  
  #Starts counting time
  time1 = Sys.time()
  
  sub_meta_temp <- sub_meta
  sub_meta_temp$condition <- sub_meta$subtype
  
  #Uses DESeq2 method to analyze the matrix data
  dds <- DESeqDataSetFromMatrix(countData = sub_matrix,
                                colData = sub_meta_temp,
                                design = ~ subject + condition)
  
  #Sets normal patients as the control
  dds$condition <- relevel(dds$condition, ref = "normal")
  
  #Prefilters to speed up method
  temp <- sub_meta |> group_by(condition) |> summarize(n=n())
  
  smallestGroupSize <- 3
  keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
  dds <- dds[keep,]
  
  #Used reduced method
  dds <- DESeq(dds, fitType = "glmGamPoi", test="LRT", reduced = ~subject)
  
  res <- results(dds)
  
  #Now the results of the genes are plotted
  
  #Gathers all plotting points for all genes
  combBC <- data.frame()
  for (gene in geneList){
    combBC <- combBC |> bind_rows(
    genePlotPoints(gene, dds, res) |> mutate(gene=gene))
  }
  
  combBC <- combBC |> mutate(log2count = log2(count))
  
  #plots the values
  graph4D <- combBC |> ggplot(aes(x=gene, y=log2count, fill=condition))+
      geom_boxplot(outliers = FALSE) + 
    labs(
      title=paste("mRNA Counts for genes of interest"),
      subtitle = "DESeq2 Subtype + Patient"
    )
  
  
  
  #Ends the time and prints time elapsed
  
  time2 = Sys.time()
  
  #Prints time elapsed
  tDif4D <- time2-time1
}
```

### Limma Analysis

```{r}
#Start time
time1 = Sys.time()

factor1 <- sub_meta$condition
subtype <- sub_meta$subtype
plate <- sub_meta$plate
subject <- sub_meta$subject


#Makes DGEList object
d0 <- DGEList(sub_matrix)

#Calculates normalization factors
d0 <- calcNormFactors(d0)

#Filter low expression genes
cutoff <- 3
drop <- which(apply(cpm(d0), 1, max) < cutoff)
d <- d0[-drop,] 

#gets lcpm data
logcpm <- cpm(d, prior.count=2, log=TRUE)

#Makes design of analysis
mm <- model.matrix(~0 + subtype + subject)

#VoomLmFit
fit <- voomLmFit(d, mm)
fit <- eBayes(fit)

cPlot = data.frame()

for( gene in geneList){
  norm <- as.data.frame(fit$EList$E[nameToID(gene), ])
  norm$subtype <- sub_meta$subtype
  norm$gene <- gene
  
  colnames(norm) <- c("counts", "subtype", "gene")
  
  cPlot <- cPlot |> bind_rows(norm)
}

cPlot$subtype <- relevel(cPlot$subtype, ref = "normal")

# #Plotting expression values
# norm <- as.data.frame(edgeR::cpm(d)[gene, ]) # to get normalized expression values.
# norm$condition <- sub_meta$condition
# colnames(norm) <- c("counts", "condition")

graph4L <- cPlot |> ggplot(aes(x = gene, y= counts, fill = subtype)) + geom_boxplot(outliers = F) + 
  labs(
    title=paste("mRNA Counts for genes of interest"),
    subtitle = "Limma-voom Subtype + Patient",
    y = "log2CPM",
  )

#ENd time
time2 = Sys.time()

tDif4L <- time2-time1
```

### Comparison

```{r}
if (runDESeq){
  graph4D
  tDif4D
}

graph4L
tDif4L
```

## Between Group Comparisons

```{r}
graph1L
graph2L
graph3L
graph4L

graph1D
graph2D
graph3D
graph4D

tDif1L
tDif2L
tDif3L
tDif4L

tDif1D
tDif2D
tDif3D
tDif4D
```
