---
title: "mRNA_NSCLC"
format: html
editor: visual
---

# Loading Libraries

```{r}
library(jsonlite)
library(tidyverse)
library(ggplot2)
library(DESeq2)
library("Biobase")
library("gplots")
library("ggplot2")
library("RColorBrewer")
library("pheatmap")
library("matrixStats")
library("lattice")
library("gtools")
library("dplyr")
library(edgeR)
library(ggsignif)
library("BiocParallel")
library("ggpubr")


```

# Data

The data used was from the TCGA database, with filters for the tissue taken from the lung, with RNA-seq data and for Gene Expression Quantification. There were 311 tsv files for normal samples and 338 NSCLC cancer samples.

When downloading data from TCGA and extracting the folders with the data, the following cmds are useful for Windows:

To move all files in subfolders into the mainfolder: (while in the parent directory)

for /r %i in (\*) do \@move "%i" .

To remove empty directories:

for /d %i in (\*) do @rmdir /s /q "%i"

```{r}
#Loads tsv files of lung cells and appends vertically
#Normal Data
files1 <- list.files("data/Normal_Lung", pattern ="*.tsv", full.names=TRUE) #Has path to folder containing files
col_name = c("gene_id","gene_name","gene_type","unstranded","stranded_first","stranded_second","tpm_unstranded","fpkm_unstranded","fpkm_uq_unstranded")
lung_normal <- read_tsv(files1, comment="#", col_names = col_name, skip=6)

#Cancer Data
files2 <- c(
  list.files("data/Cancer_Lung/Adenomas", pattern="*.tsv", full.names=TRUE), #Has path to folders containing files
  list.files("data/Cancer_Lung/Squamous", pattern="*.tsv", full.names=TRUE),
  list.files("data/Cancer_Lung/LCC", pattern="*.tsv", full.names=TRUE)
)             

lung_cancer <- read_tsv(files2, comment="#", col_names = col_name, skip=6)
```

Load the metadata

```{r}
#Load metadata and select useful columns
metadata <- fromJSON("data/Cancer_Lung/metadata.repository.lung.json")
metadata <- metadata|> bind_rows(fromJSON("data/Cancer_Lung/LCC/metadata.cart.2025-07-16.json"))

temp <- metadata |> select(associated_entities, file_name)
temp$case <- bind_rows(temp$associated_entities)$case_id
temp$sample <-sapply(temp$file_name, function(data) str_split_i(data, "\\.", 1))
meta <- temp |> select(sample, case)


#Find list of samples
normList <- sapply(files1, function(data) str_split_i(data, "/", 3))
normList <- sapply(normList, function(data) str_split_i(data, "\\.", 1))

cancList <- sapply(files2, function(data) str_split_i(data, "/", 4))
cancList <- sapply(cancList, function(data) str_split_i(data, "\\.", 1))

sampleList <- c(normList, cancList)

#Get metadata only within the list of samples
metaList <- meta[meta$sample %in% sampleList, ]
```

# Functions

## Name to ID Conversion

```{r}
nameToID <- function(geneName){
    geneID <- (unstranded_norm |> filter(gene_name==geneName))[1, ]$gene_id
  return(geneID)
}
```

## Plot Gene Normal vs Cancer

```{r}
plotGene <- function(geneName, ddsData, resData){
  geneId <- (unstranded_norm |> filter(gene_name==geneName))[1, ]$gene_id
  
  cPlot <- plotCounts(ddsData, 
                    gene=geneId, 
                    intgroup="condition", 
                    returnData = TRUE
                    )
  #Gets the adjusted p values for the gene
  pValue <- resData[geneId, ]$padj
  
  cPlot |> ggplot(aes(x=condition, y=count, fill=condition)) + geom_boxplot() + 
    labs(
      title=paste0("mRNA Counts for ", geneName),
      subtitle="Done via DESeq2 package",
      caption=paste0("p-value = ", pValue)
    )
}
```

## Plot Gene List

```{r}
plotGeneList <- function(geneList, geneName, ddsData, transformFunc = NA){
  #Makes dataframe to store list of points to plot
  pointsPlot <- data.frame()
  
  #Iterates through each gene in the list and gets the normalized expression values
  for (gene in geneList){
          
    cPlot <- plotCounts(ddsData, 
                        gene=nameToID(gene), 
                        intgroup="condition", 
                        returnData = TRUE
                        )
    cPlot <- cPlot |> mutate(gene = gene)  #Labels each gene exp with gene name
    pointsPlot <- pointsPlot |> bind_rows(cPlot)
    
  }
  
  pointsPlot$gene <- factor(pointsPlot$gene, levels = geneList)
  
  if(!is.na(transformFunc)){
    pointsPlot$trans_count <- transformFunc(pointsPlot$count)
  }
  else{
    pointsPlot$trans_count <- pointsPlot$count
  }
  
    #Makes a max limit on scale by taking highest Q3s of boxplots and extending to end of whisker
  summary <- pointsPlot |> select(trans_count, gene, condition) |> group_by(gene, condition) |> reframe(
    five = fivenum(trans_count)[4], IQR = IQR(trans_count), max = max(trans_count)
  )

  summary <- summary |> mutate(maxLim = five + 1.5*IQR)
  
  maxLim <- max(summary$maxLim)
  maxValue = max(summary$max)
  
  maxLim <- min(maxLim, maxValue)
  
  margin_top <- maxLim*0.02/maxValue
  y_pos <- maxLim + margin_top
  
  pointsPlot |> ggplot(aes(x=gene, y=trans_count, fill=condition)) + 
    ggpubr::stat_compare_means(test = 'wilcox.test', label = 'p.signif', show.legend = F, label.y= y_pos) +
    geom_boxplot(outliers = FALSE) +
    scale_x_discrete(labels = geneName) + 
    labs(
      title=paste0("Gene Expression in Non-Small Cell Lung Cancer"),
      x = "Gene Name",
      y="Gene Expression"
    ) +
    theme_classic()
  
}
```

## Plot Gene in Subtypes

```{r}
plotGeneSubtypes <- function(geneName, ddsData, name = ""){
  geneId <- (unstranded_norm |> filter(gene_name==geneName))[1, ]$gene_id
  
  cPlot <- plotCounts(ddsData, 
                    gene=geneId, 
                    intgroup="condition", 
                    returnData = TRUE
                    )
  cPlot$condition <- factor(unfactor(cPlot$condition), levels = c("normal", "adenocarcinoma", "squamous", "large cell"))
  
  #Makes a max limit on scale by taking highest Q3s of boxplots and extending to end of whisker
  summary <- cPlot |> select(count, condition) |> group_by(condition) |> reframe(
    five = fivenum(count)[4], IQR = IQR(count), max = max(count)
  )

  maxQ3 <- max(summary$five)
  temp <- (maxQ3 == summary$five)
  temp <- summary[temp, ]
  maxLim <- temp$five + 1.5*temp$IQR
  maxValue = max(summary$max)
  
  maxLim <- min(maxLim, temp$max)
  
   if(name != ""){
     geneName <- name
   }
  
  comps = list(
    c("normal", "adenocarcinoma"), 
    c("normal", "squamous"),
    c("normal", "large cell")
  )
  
  margin_top <- maxLim*0.02/maxValue
  y_pos <- maxLim + 0.08*maxLim*(0:(length(comps)-1))
  
  cPlot |> ggplot(aes(x=condition, y=count, fill=condition)) + 
    geom_boxplot(outliers= F) + 
        geom_signif(comparisons = comps, map_signif_level = TRUE, tip_length = maxLim*0.03/maxValue, y_position = y_pos) +
    labs(
      title=paste0(geneName, " Gene Expression in Non-Small Lung Cancer Subtypes"),
      y = "gene expression"
    )
}
```

## Plot Gene List in Subtypes

```{r}
plotGeneListSubtypes <- function(geneList, geneName, ddsData, transformFunc = NA){

  pointsPlot <- data.frame()
  
  for (gene in geneList){
          
    cPlot <- plotCounts(ddsData, 
                        gene=nameToID(gene), 
                        intgroup="condition", 
                        returnData = TRUE
                        )
    cPlot <- cPlot |> mutate(gene = gene) 
    pointsPlot <- pointsPlot |> bind_rows(cPlot)
    
  }
  
  pointsPlot$gene <- factor(pointsPlot$gene, levels=geneList)
  
  pointsPlot$condition <- factor(pointsPlot$condition, levels = c("normal", "adenocarcinoma", "squamous", "large cell"))
  
  condLevels <- levels(pointsPlot$condition)
  
  pointsPlot$interaction <- factor(interaction(pointsPlot$gene, pointsPlot$condition),
                           levels = c(t(sapply(condLevels, function(d) paste(geneList, d, sep=".")))))
  
  labels <- c()
  for(gene in geneList){
    labels <- c(labels, " ", gene, " ", " ")
  }
  
  #Turn genes to protein name by manually changing the names
  labels[((0:length(geneName))*4 + 2)] <- geneName
  
  
  temp <- as.data.frame(sapply(condLevels, function (d) paste(geneList, d, sep=".")))
  comps <- list()
  for(i in 1:length(geneList)){
    temp1 <- list(c(temp[i, "normal"], temp[i, "adenocarcinoma"]), c(temp[i, "normal"], temp[i, "squamous"]), c(temp[i, "normal"], temp[i, "large cell"]))
    comps <- c(comps, temp1)
  }
  
  #Applies an option transform (usually log)
  if(!is.na(transformFunc)){
    pointsPlot$trans_count <- transformFunc(pointsPlot$count)
  }
  else{
    pointsPlot$trans_count <- pointsPlot$count
  }
  
  #Makes a max limit on scale by taking highest Q3s of boxplots and extending to end of whisker
  summary <- pointsPlot |> select(trans_count, gene, condition) |> group_by(gene, condition) |> reframe(
    five = fivenum(trans_count)[4], IQR = IQR(trans_count), max = max(trans_count)
  )

  summary <- summary |> mutate(maxLim = five + 1.5*IQR)
  
  maxLim <- max(summary$maxLim)
  maxValue = max(summary$max)
  
  maxLim <- min(maxLim, maxValue)
  
  margin_top <- maxLim*0.02/maxValue
  tip_length <- maxLim*0.02/maxValue
  
  y_pos <- c(rep(maxLim + 0.08*maxLim*(0:(length(levels(pointsPlot$condition)) - 2)), length(geneList)))
  
  pointsPlot |> ggplot(aes(x=interaction, y=trans_count, fill=condition)) + 
    scale_x_discrete(
      labels = labels
    ) +
    geom_signif(comparisons = comps,
                map_signif_level  = T,
                y_position = y_pos,
                tip_length = tip_length
                ) +
    geom_boxplot(outliers=FALSE) +
    labs(
      title=paste0("Gene Expression in NSCLC Subtypes"),
      x = "Gene Name",
      y="Gene Expression"
      #caption=paste0("p-value = ", pValue)
    ) + 
  theme_classic()

}
```

# mRNA Expression Analysis

## Cleaning Data

The data from the tsv files contains unstranded counts for each gene, which will be used for the analysis. Analysis was done with the DESeq2 package. No batch correction was done since no batch data was available, so that is a possible improvement.

```{r}
#Isolates columns of interest and labels each gene by the sample it was taken from
unstranded_norm <- lung_normal |> select("gene_id", "gene_name", "unstranded") |> mutate(
  sample = ((row_number()-1) %/% 60660) + 1 #60660 different genes
)

#Turns it into a table of unstranded RNA counts
raw_norm <- unstranded_norm |> select(!gene_name) |> pivot_wider(names_from=gene_id, values_from = unstranded)
raw_norm <- raw_norm |> select(!sample)
norm_matrix_raw <- as.matrix(t(raw_norm))
colnames(norm_matrix_raw) <- paste("N", 1:dim(norm_matrix_raw)[2]
 |> lapply(toString), sep = "")

#Labels all genes with sample number
unstranded_cancer <- lung_cancer |> select("gene_id", "gene_name", "unstranded") |> mutate(
  sample = ((row_number()-1) %/% 60660) + 1
)

#Gets raw count matrix
raw_cancer <- unstranded_cancer |> select(!gene_name) |> pivot_wider(names_from=gene_id, values_from = unstranded)
raw_cancer <- raw_cancer |> select(!sample)
norm_matrix_cancer <- as.matrix(t(raw_cancer))
colnames(norm_matrix_cancer) <- paste("C", 1:dim(norm_matrix_cancer)[2] |> lapply(toString), sep = "")

#Matrices needed to work with DESeq2
#Combines cancer and normal data
norm_matrix <- cbind(norm_matrix_raw, norm_matrix_cancer)
```

The DESeq2 package requires a matrix with the gene and sample data, as well as another dataframe for all the other variables associated with the samples such as type of cells.

```{r}
#Changes column data
coldata <- data.frame(1:dim(norm_matrix)[2])
rownames(coldata) <- colnames(norm_matrix)
colnames(coldata) <- "condition"
coldata <- coldata |> mutate(
  condition = if_else(row_number() <= dim(norm_matrix_raw)[2], "normal", "cancer")
)

coldata$condition <- factor(coldata$condition)

```

Next, the analysis was done by making a DESeqDataSet using the matrix of unstranded counts and the dataframe of other variables. The pre-filter removed genes with very few counts to seep up analysis. The normal cells were set as the control and parallel cores were used for speed

## DESeq2 Analysis

```{r}
#Parallel Cores for faster running
#Need to set parallel = TRUE
library("BiocParallel")
register(SnowParam(4))

#Uses DESeq2 method to analyze the matrix data
dds <- DESeqDataSetFromMatrix(countData = norm_matrix,
                              colData = coldata,
                              design = ~ condition)
#Adding gene name data
featureData <- data.frame(gene_name=lung_normal[1:60660, "gene_name"])
mcols(dds) <- DataFrame(mcols(dds), featureData)


#Prefilters to speed up method
temp <- coldata |> group_by(condition) |> summarize(n=n())

smallestGroupSize <- min(temp$n)
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]

#Sets normal patients as the control
dds$condition <- relevel(dds$condition, ref = "normal")

#Runs all the analysis
dds <- DESeq(dds, parallel = TRUE, )
res <- results(dds)
```

The log fold shrinkage was calculated and then plotted. The resultsNames(dds) method was used to get the name for coef=

```{r}
resLFC <- lfcShrink(dds, coef="condition_cancer_vs_normal", type="apeglm", parallel = TRUE)
plotMA(resLFC, ylim=c(-10,10))
```

## Gene Expression Plots

Then, a box plot of the counts for the genes of interest. Since the data passed used the gene_id, the gene name has to be converted to gene_id in order to be used.

### Single Genes

This is the genes that are being plotted

```{r}
#To get gene name, run:
#(unstranded_norm |> filter(gene_name=="CD109"))[1, ]$gene_id

#CD97, CD109, CD151, CD166, CD276, EPCAM
plotGene("ADGRE5", dds, res)
plotGene("CD109", dds, res)
plotGene("CD151", dds, res)
plotGene("ALCAM", dds, res)
plotGene("CD276", dds, res)
plotGene("EPCAM", dds, res)



```

### Multiple Genes

The graphs are all going to be combined into one graph with all the genes

```{r}
geneList <- c("CD9", "ADGRE5", "CD109","MCAM", "CD151", "ALCAM", "CD276", "EPCAM", "FGFR4")
geneName <- c("CD9", "CD97", "CD109","CD146", "CD151", "CD166", "CD276", "EPCAM", "FGFR4")


plotGeneList(geneList[1:4], geneName[1:4], dds)

plotGeneList(geneList[5:9], geneName[5:9], dds)

plotGeneList(geneList[c(3, 6, 7, 8)], geneName[c(3, 6, 7, 8)], dds, log10)

```

# Subtype Analysis

By using separating cancer cells based on the cancer type, mRNA differences can be seen between them.

## Loading Data

```{r}
#Adenocarcinoma Data
files3 <- list.files("data/Cancer_Lung/Adenomas", pattern="*.tsv", full.names=TRUE)
col_name = c("gene_id","gene_name","gene_type","unstranded","stranded_first","stranded_second","tpm_unstranded","fpkm_unstranded","fpkm_uq_unstranded")
lung_adeno <- read_tsv(files3, comment="#", col_names = col_name, skip=6)

#Squamous Data
files4 <- list.files("data/Cancer_Lung/Squamous", pattern="*.tsv", full.names=TRUE)
lung_squamous <- read_tsv(files4, comment="#", col_names = col_name, skip=6)

files5 <- list.files("data/Cancer_Lung/LCC", pattern="*.tsv", full.names=TRUE)
lung_LCC <- read_tsv(files5, comment="#", col_names = col_name, skip=6)

```

## Cleaning Data and DESeq2 Prep

The matrix for the DESeq2 analysis is made the same way as before

```{r}

#Labels all genes with sample number
unstranded_adeno <- lung_adeno |> select("gene_id", "gene_name", "unstranded") |> mutate(
  sample = ((row_number()-1) %/% 60660) + 1
)
unstranded_squamous <- lung_squamous |> select("gene_id", "gene_name", "unstranded") |> mutate(
  sample = ((row_number()-1) %/% 60660) + 1
)

unstranded_LCC <- lung_LCC |> select("gene_id", "gene_name", "unstranded") |> mutate(
  sample = ((row_number()-1) %/% 60660) + 1
)

#Gets raw count matrix
raw_adeno <- unstranded_adeno |> select(!gene_name) |> pivot_wider(names_from=gene_id, values_from = unstranded)
raw_adeno <- raw_adeno |> select(!sample)
norm_matrix_adeno <- as.matrix(t(raw_adeno))
colnames(norm_matrix_adeno) <- paste("A", 1:dim(norm_matrix_adeno)[2] |> lapply(toString), sep = "")

raw_squamous <- unstranded_squamous |> select(!gene_name) |> pivot_wider(names_from=gene_id, values_from = unstranded)
raw_squamous <- raw_squamous |> select(!sample)
norm_matrix_squamous <- as.matrix(t(raw_squamous))
colnames(norm_matrix_squamous) <- paste("S", 1:dim(norm_matrix_squamous)[2] |> lapply(toString), sep = "")

raw_LCC <- unstranded_LCC |> select(!gene_name) |> pivot_wider(names_from=gene_id, values_from = unstranded)
raw_LCC <- raw_LCC |> select(!sample)
norm_matrix_LCC <- as.matrix(t(raw_LCC))
colnames(norm_matrix_LCC) <- paste("L", 1:dim(norm_matrix_LCC)[2] |> lapply(toString), sep = "")

#Matrices needed to work with DESeq2
#Combines cancer and normal data
norm_matrix_comb <- cbind(norm_matrix_raw, norm_matrix_adeno, norm_matrix_squamous, norm_matrix_LCC)
```

## DESeq2 Analysis

This runs the DESeq2 analysis

```{r}
#Changes column data
coldataComb <- data.frame(1:dim(norm_matrix_comb)[2])
rownames(coldataComb) <- colnames(norm_matrix_comb)
colnames(coldataComb) <- "condition"
coldataComb <- coldataComb |> mutate(
  condition = if_else(row_number() <= dim(norm_matrix_raw)[2], "normal", 
                      if_else(row_number() <= dim(norm_matrix_raw)[2] + dim(norm_matrix_adeno)[2], "adenocarcinoma", 
                              if_else(row_number() <= dim(norm_matrix_raw)[2] + dim(norm_matrix_adeno)[2] + dim(norm_matrix_squamous)[2], "squamous", "large cell")
                              )
                      )
)

coldataComb$condition <- factor(coldataComb$condition)
coldataComb$sample <- factor(metaList$sample)
coldataComb$case <- factor(metaList$case)
```

This sets up and runs the DESeq2 analysis.

```{r}
#Parallel Cores for faster running
#Need to set parallel = TRUE
library("BiocParallel")
register(SnowParam(4))

#Uses DESeq2 method to analyze the matrix data
dds2 <- DESeqDataSetFromMatrix(countData = norm_matrix_comb,
                              colData = coldataComb,
                              design = ~ condition)
#Adding gene name data
featureData <- data.frame(gene_name=lung_normal[1:60660, "gene_name"])
mcols(dds2) <- DataFrame(mcols(dds2), featureData)


#Prefilters to speed up method
temp <- coldataComb |> group_by(condition) |> summarize(n=n())

smallestGroupSize <- min(temp$n)
keep <- rowSums(counts(dds2) >= 10) >= smallestGroupSize
dds2 <- dds2[keep,]

#Sets normal patients as the control
dds2$condition <- relevel(dds2$condition, ref = "normal")

#Runs all the analysis
dds2 <- DESeq(dds2, parallel = TRUE)
resSquamous <- results(dds2, contrast = c("condition", "squamous", "normal"))
resAdeno <- results(dds2, contrast = c("condition", "adenocarcinoma", "normal"))
resLCC <- results(dds2, contrast = c("condition", "large cell", "normal"))
```

## Plotting Expression Data

### Single Gene

Plots

```{r}

#To get gene name, run:
#(unstranded_norm |> filter(gene_name=="CD109"))[1, ]$gene_id
#CD151
resList <- list(resAdeno, resSquamous)



plotGeneSubtypes("ADGRE5", dds2, "CD109")
ggsave("plots/CD109_NSCLC_gene_expression.png")

plotGeneSubtypes("CD151", dds2)
ggsave("plots/CD151_NSCLC_gene_expression.png")

plotGeneSubtypes("ALCAM", dds2, "CD166")
ggsave("plots/CD166_NSCLC_gene_expression.png")

plotGeneSubtypes("EPCAM", dds2)
ggsave("plots/EPCAM_NSCLC_gene_expression.png")

plotGeneSubtypes("CD276", dds2)
ggsave("plots/CD276_NSCLC_gene_expression.png")

```

### Multiple Genes

Now to combine all the graphs

```{r}

plotGeneListSubtypes(geneList[1:4], geneName[1:4], dds2)
plotGeneListSubtypes(geneList[5:9], geneName[5:9], dds2)
plotGeneListSubtypes(geneList[c(3, 6, 7, 8)], geneName[c(3, 6, 7, 8)], dds2, log10)

```

# Limma Analysis

Function

```{r}
#Converts a gene name to an id
nameToID <- function(geneName){
  geneId <- (unstranded_norm |> filter(gene_name==geneName))[1, ]$gene_id
  geneId <- as.vector(geneId)[1]
  return(geneId)
}
```

Limma

```{r}
#Make metadata
condition <- coldataComb$condition
case <- coldataComb$case


condition <- relevel(condition, ref="normal")

#Creates DGEList object
d0 <- DGEList(norm_matrix_comb)

#Calculates normalization factors
d0 <- calcNormFactors(d0)

#Filter low expression genes
cutoff <- 100
drop <- which(apply(cpm(d0), 1, max) < cutoff)
d <- d0[-drop,] 

#MDS Plot
#plotMDS(d, col = as.numeric(condition))

#LogCPM data
logcpm <- cpm(d, prior.count=2, log=TRUE)


#Makes design of analysis
mm <- model.matrix(~0 + condition)

#VoomLmFit
fit <- voomLmFit(d, mm, plot = T)
fit <- eBayes(fit)

cPlot = data.frame()

for( gene in geneList){
  norm <- as.data.frame(fit$EList$E[nameToID(gene), ])
  norm$condition <- coldataComb$condition
  norm$gene <- gene
  
  colnames(norm) <- c("counts", "condition", "gene")
  
  cPlot <- cPlot |> bind_rows(norm)
}

# #Plotting expression values
# norm <- as.data.frame(edgeR::cpm(d)[gene, ]) # to get normalized expression values.
# norm$condition <- sub_meta$condition
# colnames(norm) <- c("counts", "condition")

cPlot$condition <- relevel(cPlot$condition, ref = "normal")

cPlot$gene <- factor(cPlot$gene, levels = geneList)

cPlot |> ggplot(aes(x = gene, y= counts, fill = condition)) + geom_boxplot(outliers = F) + 
  labs(
    title=paste("mRNA Counts for genes of interest"),
    subtitle = "Limma-voom",
    y = "log2CPM"
  )
```

# Quality Analysis

## Heatmap

This produces a heat map of the normal vs cancer dataset

```{r}
library("pheatmap")
ntd <- normTransform(dds2)
vsd <- vst(dds2, blind=FALSE)

select <- order(rowMeans(counts(dds2,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds2)[,c("condition")])
rownames(df) <- colnames(assay(ntd)[select,])
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)

```

Heatmap

```{r}
select <- order(rowMeans(counts(dds2,normalized=TRUE)),
                decreasing=TRUE)[1:20]
df <- as.data.frame(colData(dds2)[,c("condition")])
rownames(df) <- colnames(assay(ntd)[select,])
pheatmap(assay(ntd)[select,], cluster_rows=FALSE, show_rownames=FALSE, cluster_cols=FALSE, annotation_col=df)

```

Sample to Sample diffs

```{r}
sampleDists <- dist(t(assay(vsd)))

sampleDistMatrix <- as.matrix(sampleDists)
rownames(sampleDistMatrix) <- paste(vsd$condition, vsd$type, sep="-")
colnames(sampleDistMatrix) <- NULL
colors <- colorRampPalette( rev(brewer.pal(9, "Blues")) )(255)
pheatmap(sampleDistMatrix,
         clustering_distance_rows=sampleDists,
         clustering_distance_cols=sampleDists,
         col=colors)


```

```{r}
#plotPCA(vsd, intgroup=c("condition", "type"))

```
