---
title: "ProteomicDB"
format: html
editor: visual
---

# Loading Libraries

```{r}
#| include: false

library(jsonlite)
library(httr) #For REST API calls
library(tidyverse)
library(ggplot2) #Plotting
library(DESeq2) #DE Analysis
library("Biobase") 
library("gplots")
library("RColorBrewer") 
library("pheatmap") #For heat maps and quality control graphs
library("matrixStats")
library("lattice")
library("gtools")
library("dplyr")
library(edgeR) #For data analysis and limma
library("BiocParallel") #For running DESeq2 in parallel
library(ggsignif)
library("ggpubr") #Adding significance bars for graphs
library(limma) #For removing batch effects???
```

# Loading Data

```{r}
#Data downloaded from proteomicsDB as the lung expression data
CD97_1 <- read_delim("data/ProteomicsDB/CD97_P48960.csv")  

```

# Functions

## API Call

Function to get data from an API Call to proteonomicsDB

API Documentation: <https://www.proteomicsdb.org/apiv2>

Returns either a list or dataframe containing the necessary data

```{r}
getAPIData <- function(uriString){
  #Turns strings and other special characters into percent encoding to be URL compatible
  uriSafe <- URLencode(uriString, reserved = FALSE)
  uriSafe <- str_replace_all(uriSafe, coll("'"), "%27")
  uriSafe <- str_replace_all(uriSafe, coll("("), "%28")
  uriSafe <- str_replace_all(uriSafe, coll(")"), "%29")
  uriSafe <- str_replace_all(uriSafe, coll(" "), "%20")
  uriSafe <- paste0(substr(uriSafe, start = 1, stop = 8), 
                    str_replace_all(
    substr(uriSafe, start = 9, stop = str_length(uriSafe)),
    coll(":"), "%3A"
    )
  )

  #print(uriSafe)
  
  #Get the data from the API call
  res = GET(uriSafe)
  
  #Get status code
  status <- res$status_code
  
  #If API call succeeds
  if(status == 200) {
    #Calls data
    data <- fromJSON(rawToChar(res$content))$d
    
    #Checks if the returned data has an extra layer
    if(length(names(data)) == 1)
      return(data$results)
    else
      return(data)
  } else {
    print(paste("API CALL FAILED. ERROR CODE", status))
  }
}

```

## Protein Name Data

Function to view data associated with a gene name

```{r}
viewGeneData <- function(gene_name){
  #URI to API call
  url <- paste0(
    "https://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Protein?$format=json&$filter=GENE_NAME eq '",
    gene_name, 
    "' and DECOY eq 0")
  
  #Replace special characters like " " and "'" with url friendly versions (%20 and %27)
  data <- getAPIData(url)
  View(data)
}
```

## Name to ID

Function to convert the protein name to the corresponding protein id, as well as the ids for its isoforms

```{r}
proteinNameToID <- function(gene_name){
  #makes url for API Call
    url <- paste0(
    "https://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Protein?$format=json&$filter=ENTRY_NAME eq '",
    gene_name, "_HUMAN",
    "' and DECOY eq 0")
    
    #Calls API
    data <- getAPIData(url)
    
    #Returns protein ID for the protein and its isoforms
    #1st element is canonical form, next entries are the isoforms
    return(data$PROTEIN_ID)
}
```

## Lung

### Single Protein Lung

Function to plot or return data associated with a specific protein in lung samples

Takes the name of protein, dataframes of normal and cancer lung sample data and returns the expression to plot

```{r}
pointsProteinLung <- function(protein, lung_normal_data, lung_cancer_data){
  #Allows the function to work with protein names or ids
  if(class(protein)=="character"){
    protein_name <- protein
    protein_id <- proteinNameToID(protein_name)[1]
  } else {
    protein_id <- protein
  }
  
  #Iterates through each sample's protein expression data
  lung_normal_exp_data <- list()
  
  #Takes each protein expression API call from each sample and filters for the specific protein
  for(normLink in lung_normal_data$ProteinExpression$`__deferred`$uri){
    temp <- list(getAPIData(paste0(normLink, "?$format=json&$filter=PROTEIN_ID eq ", protein_id)))
    lung_normal_exp_data <- c(lung_normal_exp_data, temp) #List of exp datas
  }
  
  #Combines all the sample expression data together
  lung_normal_exp_CD97 <- bind_rows(lung_normal_exp_data)
  
  #Iterates through each sample's protein expression data
  lung_cancer_exp_data <- list()
  for(cancLink in lung_cancer_data$ProteinExpression$`__deferred`$uri){
    temp <- list(getAPIData(paste0(cancLink, "?$format=json&$filter=PROTEIN_ID eq ", protein_id)))
    lung_cancer_exp_data <- c(lung_cancer_exp_data, temp)
  }
  
  #Joins all cancer exp data into a data frame
  lung_cancer_exp_CD97_data <- bind_rows(lung_cancer_exp_data) 
  
  #Checks the rest of the list
  canc_data_list <- list()
  
  #Iterates through all other cancer tissues
  for (tissueType in lung_sample_data[-2]){
    lung_cancer_exp_data_2 <- list()
    
    #Gets protein expression for each cancer tissue type
    for(cancLink in tissueType$ProteinExpression$`__deferred`$uri){
      #Get API data
      temp <- list(getAPIData(paste0(cancLink, 
                                     "?$format=json&$filter=PROTEIN_ID eq ", protein_id)))
      lung_cancer_exp_data_2 <- c(lung_cancer_exp_data_2, temp)
    }
     #Gets all samples for a specific tissue type and combines into a single data frame
    lung_cancer_exp_CD97_data_2 <- bind_rows(lung_cancer_exp_data_2)
    
    #Adds the dataframe to the list of samples for each tissue type
    canc_data_list <- c(canc_data_list, list(lung_cancer_exp_CD97_data_2))
  }
  
  #Combine cancer expression dataframes
  lung_cancer_exp_CD97 <- lung_cancer_exp_CD97_data |>
    bind_rows(
      bind_rows(canc_data_list)
    )
  
  #Sets up the points to plot
  #Choose which set of protein expression to use
  normal_exp <- lung_normal_exp_CD97 |> filter(CALCULATION_METHOD == calc_method) |> select(EXPRESSION) |> mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  cancer_exp <- lung_cancer_exp_CD97 |> filter(CALCULATION_METHOD == calc_method) |> select(EXPRESSION) |> mutate(condition = "cancer", expression = as.numeric(EXPRESSION))
  
  #Combine normal + cancer data
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  #Sets levels
  plotData$condition <- factor(plotData$condition, levels = c("normal", "cancer"))
  
  return(plotData)
}
```

Function to plot the points given in the previous section

```{r}
plotProteinLung <- function(protein_name, lung_normal_data, lung_cancer_data){
  
  
  #Gets data to plot the ponts
  plotData <- pointsProteinLung(protein_name, lung_normal_data, lung_cancer_data)
  
  #Get number of normal and cancer data points
  nNormal <- dim(plotData |> filter(condition == "normal"))[1]
  nCancer <- dim(plotData |> filter(condition == "cancer"))[1]
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = condition, y = expression, fill = condition)) +
  geom_boxplot() + 
  labs(
    title = paste("Protein Expression of", protein_name, " in Lung Cancers"),
    subtitle = paste0(nNormal, " normal samples, ", nCancer, " cancer samples"),
  ) + 
  stat_compare_means(label = "p.signif", comparisons = list(c("normal", "cancer")))
  
}
```

### Multiple Protein Lung

Return points to plot for a series of protein names

```{r}
pointsProteinListLung <- function(protein_list, lung_normal_data, lung_cancer_data){
  #Sets up the points to plot
  
  #Gets list of protein ids to use in api call
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1]) #1st element is protein
  
  #Iterates through each sample's protein expression data
  lung_normal_exp_data <- list()
  for(normLink in lung_normal_data$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets the API Call
    temp <- list(getAPIData(uri))
    lung_normal_exp_data <- c(lung_normal_exp_data, temp)
  }
  
  #Combines all normal sample data
  lung_normal_exp <- bind_rows(lung_normal_exp_data)
  
  #Selects columns of interest and labels data
  normal_exp <- lung_normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #Cancer Data
  #Iterates through each sample's protein expression data
  lung_cancer_exp_data <- list()
  for(cancLink in lung_cancer_data$ProteinExpression$`__deferred`$uri){
    #First part of API Call
    uri <- paste0(cancLink, "?$format=json&$filter=")
    
    #Joins additional filters for each protein id
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets sample data and combines with other sample data
    temp <- list(getAPIData(uri))
    lung_cancer_exp_data <- c(lung_cancer_exp_data, temp)
  }
  #Combines all sample data into dataframe
  lung_cancer_exp <- bind_rows(lung_cancer_exp_data)
  
  
  #Checks the rest of the list
  canc_data_list <- list()
  
  #Looks through the other list elements
  for (tissueType in lung_sample_data[-2]){
    lung_cancer_exp_data_2 <- list()
    
    #Ensures that the lung data isn't treated with anything
    tissueType <- tissueType |> filter(TREATMENT %in% c("Not applicable/none", ""))
    
    #Gets data from each cancer sample
    for(cancLink in tissueType$ProteinExpression$`__deferred`$uri){
      uri <- paste0(cancLink, "?$format=json&$filter=")
      temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
      temp <- paste(temp, collapse = " or ")
      uri <- paste0(uri, temp)
    
      
      temp <- list(getAPIData(uri))
      lung_cancer_exp_data_2 <- c(lung_cancer_exp_data_2, temp)
    }
    
    #Combines all cancer sample data
    lung_cancer_exp_data_2 <- bind_rows(lung_cancer_exp_data_2)
    
    canc_data_list <- c(canc_data_list, list(lung_cancer_exp_data_2))
  }
  
  #Combines all the cancer data together into single dataframe
  lung_cancer_exp <- lung_cancer_exp |>
    bind_rows(
      bind_rows(canc_data_list)
    )
  
  #Gets only columns of interest
  cancer_exp <- lung_cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "cancer", expression = as.numeric(EXPRESSION))
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  #Factors columns
  plotData$condition <- factor(plotData$condition, levels = c("normal", "cancer"))
  plotData$PROTEIN_ID <- factor(plotData$PROTEIN_ID)
  
  return(plotData)
}
```

Plotting Function

```{r}
plotProteinListLung <- function(protein_list, lung_normal_data, lung_cancer_data, protein_list_name){
  
  #Gets data to plot the ponts
  plotData <- pointsProteinListLung(protein_list, lung_normal_data, lung_cancer_data)
  
  #Gets list of ids for labeling data with protein names
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_list_name[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list_name)
  
  #Gets number of samples in each pair of proteins and conditions
  n <- plotData |> 
    group_by(protein_name, condition) |> 
    summarize(n = n()) |> 
    pivot_wider(names_from=condition, values_from = n) |>
    mutate(label = paste0(normal,"N\n", cancer,"C"))

  #Get the number of proteins and sample numbers for annotations
  nProt <- length(protein_list_name)
  labels <- n$label
  
  #Only compare proteins with normal and cancer samples
  valid_list <- c()
  for(index in 1:length(protein_list_name)){
    protein <- protein_list_name[index]
  
    plotData_protein <- plotData |> filter(protein_name == protein)
    levelsPlot <- unique(plotData_protein$condition)
    if(length(levelsPlot) == 2){
      valid_list <- c(valid_list, protein)
    }
  }
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = protein_name, y = expression, fill = condition)) +
    geom_boxplot(outliers = F) + 
    labs(
    title = paste("Protein Expression in Non-Small Cell Lung Cancer"),
    x = "Protein Name",
    y = "Protein Expression"
  ) + 
    stat_compare_means(data = plotData |> filter(protein_name %in% valid_list), label = "p.signif") + #Statistical significance
    theme_classic()
}
```

### Combining Isoforms

This gets the protein expression points for the proteins and their isoforms for a list of protein names

```{r}
pointsProteinListLungIso <- function(protein_list, lung_normal_data, lung_cancer_data){
  #Sets up the points to plot
  #Choose which set of protein expression to use

  #Gets list of protein ids to use in api call from the proteins and their isoforms
  protein_id_list <- c()
  protein_name_list <- c()
  for(protein in protein_list)
  {
    iso_list <- proteinNameToID(protein)
    protein_id_list <- c(protein_id_list, iso_list)
    protein_name_list <- c(protein_name_list, rep(protein, length(iso_list)))
  }
  
  
  #Iterates through each sample's protein expression data
  lung_normal_exp_data <- list()
  for(normLink in lung_normal_data$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets API Call
    temp <- list(getAPIData(uri))
    lung_normal_exp_data <- c(lung_normal_exp_data, temp)
  }
  
  lung_normal_exp <- bind_rows(lung_normal_exp_data)
  
  #Filters for useful columns and levels data
  normal_exp <- lung_normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #Cancer Data
  #Iterates through each sample's protein expression data
  lung_cancer_exp_data <- list()
  for(cancLink in lung_cancer_data$ProteinExpression$`__deferred`$uri){
    #Builds url for API call
    uri <- paste0(cancLink, "?$format=json&$filter=")
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets and combines API call data
    temp <- list(getAPIData(uri))
    lung_cancer_exp_data <- c(lung_cancer_exp_data, temp)
  }
  
  #Combines cancer data into a dataframe
  lung_cancer_exp <- bind_rows(lung_cancer_exp_data)
  
  
  #Checks the rest of the list
  canc_data_list <- list()
  
  #Looks through the other list elements
  for (tissueType in lung_sample_data[-2]){
    lung_cancer_exp_data_2 <- list()
    
    for(cancLink in tissueType$ProteinExpression$`__deferred`$uri){
      uri <- paste0(cancLink, "?$format=json&$filter=")
      temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
      temp <- paste(temp, collapse = " or ")
      uri <- paste0(uri, temp)
    
      temp <- list(getAPIData(uri))
      lung_cancer_exp_data_2 <- c(lung_cancer_exp_data_2, temp)
    }
    
    lung_cancer_exp_data_2 <- bind_rows(lung_cancer_exp_data_2)
    
    canc_data_list <- c(canc_data_list, list(lung_cancer_exp_data_2))
  }
  
  #Combines all cancer data into single dataframe
  lung_cancer_exp <- lung_cancer_exp |>
    bind_rows(
      bind_rows(canc_data_list)
    )
  
  #Filters for useful columns
  cancer_exp <- lung_cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "cancer", expression = as.numeric(EXPRESSION))
  
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  plotData$condition <- factor(plotData$condition, levels = c("normal", "cancer"))
  plotData$PROTEIN_ID <- factor(plotData$PROTEIN_ID)
  
  return(plotData)
}
```

Plotting Function

```{r}
plotProteinListLungIso <- function(protein_list, lung_normal_data, lung_cancer_data){
  
  #Gets data to plot the ponts
  plotData <- pointsProteinListLungIso(protein_list, lung_normal_data, lung_cancer_data)
  
  #Gets list of protein ids to use in api call
  protein_id_list <- c()
  protein_name_list <- c()
  for(protein in protein_list)
  {
    iso_list <- proteinNameToID(protein)
    protein_id_list <- c(protein_id_list, iso_list)
    protein_name_list <- c(protein_name_list, rep(protein, length(iso_list)))
  }
  
  #Labels each data point with the protein name
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_name_list[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list)
  
  #Gets number of data points for each pair of proteins and conditions
  n <- plotData |> 
    group_by(protein_name, condition) |> 
    summarize(n = n()) |> 
    pivot_wider(names_from=condition, values_from = n) |>
    mutate(label = paste0(normal,"N\n", cancer,"C"))

  #Gets data for annotations
  nProt <- length(protein_list)
  labels <- n$label
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = protein_name, y = expression, fill = condition)) +
    geom_boxplot() + 
    labs(
    title = paste("Protein Expression with Isoforms in Lung Cancers"),
  ) + 
    stat_compare_means(label = "p.signif") +
    annotate("text", x=1:nProt, y=4, label=labels)
  
}
```

### Subtype Single

```{r}
pointsProteinLungSubtype <- function(protein_name, lung_normal_data, lung_cancer_list, subtype_list){
  
  protein_id <- proteinNameToID(protein_name)[1]
  
  #Iterates through each sample's protein expression data
  lung_normal_exp_data <- list()
  
  #Takes each protein expression API call from each sample and filters for the specific protein
  for(normLink in lung_normal_data$ProteinExpression$`__deferred`$uri){
    temp <- list(getAPIData(paste0(normLink, "?$format=json&$filter=PROTEIN_ID eq ", protein_id)))
    lung_normal_exp_data <- c(lung_normal_exp_data, temp) #List of exp datas
  }
  
  #Combines all the sample expression data together
  lung_normal_exp <- bind_rows(lung_normal_exp_data)
  #View(lung_normal_exp |> filter(CALCULATION_METHOD == calc_method))
  
  
  #Iterates through each sample's protein expression data
  lung_cancer_exp <- data.frame()
  
  for(i in 1:length(subtype_list)){
    subtype <- subtype_list[i]
    sample_cancer_data <- lung_cancer_list[[i]]
    
    lung_subtype_exp <- data.frame()

    for(cancLink in sample_cancer_data$ProteinExpression$`__deferred`$uri){
      temp <- as.data.frame(getAPIData(paste0(cancLink, "?$format=json&$filter=PROTEIN_ID eq ", protein_id)))
      lung_subtype_exp <- bind_rows(lung_subtype_exp, temp)
    }
    
    lung_subtype_exp <- lung_subtype_exp |> mutate(condition = subtype)
    
    lung_cancer_exp <- lung_cancer_exp |> bind_rows(lung_subtype_exp)
  }
  #View(lung_cancer_exp |> filter(CALCULATION_METHOD == calc_method))

  
  #Sets up the points to plot
  #Choose which set of protein expression to use
  normal_exp <- lung_normal_exp |> filter(CALCULATION_METHOD == calc_method) |> select(EXPRESSION, SAMPLE_ID) |> mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  cancer_exp <- lung_cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, condition, SAMPLE_ID) |> mutate(expression = as.numeric(EXPRESSION))
  
  #Combine normal + cancer data
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  #Sets levels
  plotData$condition <- factor(plotData$condition, levels = c("normal", "adeno", "squamous", "large cell"))
  
  levels(plotData$condition) <- c("normal", "adenocarcinoma", "squamous", "large cell")
  
  return(plotData)
}
```

Plotting

```{r}
plotProteinLungSubtype <- function(protein_name, lung_normal_data, lung_cancer_list, subtype_list){
  
  plotData <- pointsProteinLungSubtype(protein_name, lung_normal_data, lung_cancer_list, subtype_list)
  
  comps <- list(
    c("normal", "adenocarcinoma"),
    c("normal", "squamous"),
    c("normal", "large cell")
  )

  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = condition, y = expression, fill = condition)) +
    geom_boxplot(outliers = F) + 
    labs(
    title = paste(protein_name, "Protein Expression in Non-Small Lung Cancer Subtypes"),
    x = "condition",
    y = "protein expression"
  ) + 
    geom_signif(
      comparisons = comps,
      map_signif_level  = T,
      step_increase = 0.08,
      test.args = list(paired = F)
    ) 
}
```

### Subtype List

Points

```{r}
pointsProteinListLungSubtype <- function(protein_list, lung_normal_data, lung_cancer_list, subtype_list){
  
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  #Iterates through each sample's protein expression data
  lung_normal_exp_data <- list()
  
  #Takes each protein expression API call from each sample and filters for the specific protein
  for(normLink in lung_normal_data$ProteinExpression$`__deferred`$uri){
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    temp <- list(getAPIData(uri))
    lung_normal_exp_data <- c(lung_normal_exp_data, temp) #List of exp datas
  }
  
  #Combines all the sample expression data together
  lung_normal_exp <- bind_rows(lung_normal_exp_data)
  #View(lung_normal_exp |> filter(CALCULATION_METHOD == calc_method))
  
  
  #Iterates through each sample's protein expression data
  lung_cancer_exp <- data.frame()
  
  for(i in 1:length(subtype_list)){
    subtype <- subtype_list[i]
    sample_cancer_data <- lung_cancer_list[[i]]
    
    lung_subtype_exp <- data.frame()

    for(cancLink in sample_cancer_data$ProteinExpression$`__deferred`$uri){
      uri <- paste0(cancLink, "?$format=json&$filter=")
      #Joins filters for each protein ID
      temp <- sapply(protein_id_list, 
                     function(data) paste0("PROTEIN_ID eq ", data)
                     )
      #Joins them together
      temp <- paste(temp, collapse = " or ")
      uri <- paste0(uri, temp)
      
      temp <- as.data.frame(getAPIData(uri))
      lung_subtype_exp <- bind_rows(lung_subtype_exp, temp)
    }
    
    lung_subtype_exp <- lung_subtype_exp |> mutate(condition = subtype)
    
    lung_cancer_exp <- lung_cancer_exp |> bind_rows(lung_subtype_exp)
  }
  #View(lung_cancer_exp |> filter(CALCULATION_METHOD == calc_method))

  
  #Sets up the points to plot
  #Choose which set of protein expression to use
  normal_exp <- lung_normal_exp |> filter(CALCULATION_METHOD == calc_method) |> select(EXPRESSION, PROTEIN_ID, SAMPLE_ID) |> mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  cancer_exp <- lung_cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION,  PROTEIN_ID, condition, SAMPLE_ID) |> mutate(expression = as.numeric(EXPRESSION))
  
  #Combine normal + cancer data
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  #Sets levels
  plotData$condition <- factor(plotData$condition, levels = c("normal", "adeno", "squamous", "large cell"))
  
  levels(plotData$condition) <- c("normal", "adenocarcinoma", "squamous", "large cell")
  
  return(plotData)
}
```

Plot

```{r}
plotProteinListLungSubtype <- function(protein_list, lung_normal_data, lung_cancer_list, subtype_list, protein_list_name){
  
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  plotData <- pointsProteinListLungSubtype(protein_list, lung_normal_data, lung_cancer_list, subtype_list)
  
  
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_list_name[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list_name)
  
  
  #Gets different conditions in proper order
  condLevels <- levels(plotData$condition)
  
  #Makes an interaction that gives every combination between protein and condition to label each box plot
  plotData$interaction <- factor(
    interaction(plotData$protein_name, plotData$condition), 
    levels = c(t(sapply(condLevels, function(d) paste(protein_list_name, d, sep="."))))#Orders the factor levels for boxplot
    )
  
  #Generates labels for the x axis by making a central boxplot have the protein name and keeping the others blank
  xlabels <- c()
  for(protein in protein_list_name){
    xlabels <- c(xlabels, " ", protein, " ", " ")
  }
  
  #Creates list of comparisons to make by for each protein, getting the comparisons to be made
  temp <- as.data.frame(sapply(condLevels, function (d) paste(protein_list_name, d, sep=".")))
  comps <- list()
  for(i in 1:length(protein_list_name)){
    temp1 <- list(c(temp[i, "normal"], temp[i, "adenocarcinoma" ]), 
                  c(temp[i, "normal"], temp[i, "squamous" ]),
                  c(temp[i, "normal"], temp[i, "large cell" ])
                  )
    comps <- c(comps, temp1)
  }

  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = interaction, y = expression, fill = condition)) +
    scale_x_discrete(
      labels = xlabels 
    ) +
    geom_boxplot(outliers = F) + 
    labs(
    title = paste("Protein Expression in NSCLC Subtypes"),
    x = "Protein Name",
    y="Protein Expression"
  ) + 
    geom_signif(
      comparisons = comps,
      map_signif_level  = T,
      y_position = c(rep(c(10, 10.5, 11), 6))
    ) + 
    theme_classic()
  
}
```

## Breast

### Multiple Proteins

Takes a list of proteins (can be a single protein) along with the data frame of normal and cancerous breast cancer samples and returns protein expressions to plot

```{r}
pointsProteinList <- function(protein_list, normal_samples, cancer_samples){
  #Sets up the points to plot
  
  #Gets list of protein ids to use in api call
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  #count <- 1
  
  #Iterates through each sample's protein expression data
  normal_exp_data <- list()
  for(normLink in normal_samples$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets API Call
    temp <- list(getAPIData(uri))
    normal_exp_data <- c(normal_exp_data, temp)
    
    #count <- count +1
    #print(count) #FOR DEBUGGING
  }
  
  normal_exp <- bind_rows(normal_exp_data)
  
  #Filters for useful columns
  normal_exp <- normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID, CALCULATION_METHOD) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #Cancer Data
  #Iterates through each sample's protein expression data
  cancer_exp_data <- list()
  for(cancLink in cancer_samples$ProteinExpression$`__deferred`$uri){
    uri <- paste0(cancLink, "?$format=json&$filter=")
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    temp <- list(getAPIData(uri))
    cancer_exp_data <- c(cancer_exp_data, temp)
    
    #count <- count + 1
    #print(count) #FOR DEBUGGING
  }
  
  #Combines cancer data and filters for useful columns
  cancer_exp <- bind_rows(cancer_exp_data)
  
  cancer_exp <- cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID, CALCULATION_METHOD) |> 
    mutate(condition = "cancer", expression = as.numeric(EXPRESSION))
  
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  plotData$condition <- factor(plotData$condition, levels = c("normal", "cancer"))
  plotData$PROTEIN_ID <- factor(plotData$PROTEIN_ID)
  
  return(plotData)
}
```

Plotting

```{r}
plotProteinList <- function(protein_list, normal_samples, cancer_samples, protein_list_name){
  
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  #Gets data to plot the ponts
  plotData <- pointsProteinList(protein_list, normal_samples, cancer_samples)
  
  #Gets list of protein ids and labels each data point with protein name
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_list_name[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list_name)
  
  #Gets number of data points for each pair of proteins and conditions
  n <- plotData |> 
    group_by(protein_name, condition) |> 
    summarize(n = n()) |> 
    pivot_wider(names_from=condition, values_from = n) |>
    mutate(label = paste0(normal,"N\n", cancer,"C"))

  nProt <- length(protein_list)
  labels <- n$label
  
  #Only compare proteins with normal and cancer samples
  valid_list <- c()
  for(index in 1:length(protein_list_name)){
    protein <- protein_list_name[index]
  
    plotData_protein <- plotData |> filter(protein_name == protein)
    levelsPlot <- unique(plotData_protein$condition)
    if(length(levelsPlot) == 2){
      valid_list <- c(valid_list, protein)
    }
  }
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = protein_name, y = expression, fill = condition)) +
    geom_boxplot(outliers = F) + 
    labs(
    title = paste("Protein Expression in Breast Cancer"),
    x = "Protein Name",
    y = "Protein Expression"
  ) + 
    stat_compare_means(data = plotData |> filter(protein_name %in% valid_list), test = "wilcox.test", label = "p.signif") + 
    theme_classic()
}
```

### Isoforms Combine

```{r}
pointsProteinListIso <- function(protein_list, normal_samples, cancer_samples){
  #Sets up the points to plot
  
  #Gets list of protein ids to use in api call
  protein_id_list <- c()
  protein_name_list <- c()
  for(protein in protein_list)
  {
    iso_list <- proteinNameToID(protein)
    protein_id_list <- c(protein_id_list, iso_list)
    protein_name_list <- c(protein_name_list, rep(protein, length(iso_list)))
  }
  
  #Iterates through each sample's protein expression data
  normal_exp_data <- list()
  for(normLink in normal_samples$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets API Call
    temp <- list(getAPIData(uri))
    normal_exp_data <- c(normal_exp_data, temp)
  }
  
  normal_exp <- bind_rows(normal_exp_data)
  
  
  normal_exp <- normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #Cancer Data
  #Iterates through each sample's protein expression data
  cancer_exp_data <- list()
  for(cancLink in cancer_samples$ProteinExpression$`__deferred`$uri){
    uri <- paste0(cancLink, "?$format=json&$filter=")
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    temp <- list(getAPIData(uri))
    cancer_exp_data <- c(cancer_exp_data, temp)
  }
  
  cancer_exp <- bind_rows(cancer_exp_data)
  
  cancer_exp <- cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "cancer", expression = as.numeric(EXPRESSION))
  
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  plotData$condition <- factor(plotData$condition, levels = c("normal", "cancer"))
  plotData$PROTEIN_ID <- factor(plotData$PROTEIN_ID)
  
  return(plotData)
}
```

Plotting

```{r}
plotProteinListIso <- function(protein_list, normal_samples, cancer_samples){
  
  #Gets data to plot the ponts
  plotData <- pointsProteinListIso(protein_list, normal_samples, cancer_samples)
  
  #Gets list of protein ids to use in api call
  protein_id_list <- c()
  protein_name_list <- c()
  for(protein in protein_list)
  {
    iso_list <- proteinNameToID(protein)
    protein_id_list <- c(protein_id_list, iso_list)
    protein_name_list <- c(protein_name_list, rep(protein, length(iso_list)))
  }
  
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_name_list[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list)
  
  n <- plotData |> 
    group_by(protein_name, condition) |> 
    summarize(n = n()) |> 
    pivot_wider(names_from=condition, values_from = n) |>
    mutate(label = paste0(normal,"N\n", cancer,"C"))

  nProt <- length(protein_list)
  labels <- n$label
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = protein_name, y = expression, fill = condition)) +
    geom_boxplot() + 
    labs(
    title = paste("Protein Expression with Isoforms in Breast Cancers"),
  ) + 
    stat_compare_means(label = "p.signif") +
    annotate("text", x=1:nProt, y=4, label=labels)
  
}
```

### Subtype Single Protein

Points

```{r}
pointsProteinSubtype <- function(protein_name, normal_samples, cancer_samples){
  #Sets up the points to plot
  
  #Gets list of protein ids to use in api call
  protein_id <- proteinNameToID(protein_name)[1]
  
  #Iterates through each sample's protein expression data
  normal_exp_data <- list()
  for(normLink in normal_samples$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=PROTEIN_ID eq ", protein_id)
    
    #Gets API Call
    temp <- list(getAPIData(uri))
    normal_exp_data <- c(normal_exp_data, temp)
  }
  
  #Combines all the expression data together into a single data frame
  normal_exp <- bind_rows(normal_exp_data)
  
  #Filters out only exp data from a specific calculation method and only selects relevant columns
  normal_exp <- normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #Cancer Data
  #Iterates through each sample's protein expression data
  cancer_exp_data <- list()
  
  nCanc <- dim(cancer_samples)[1]
  
  #Iterates through each cancer sample
  for(index in 1:nCanc){
    cancLink <- cancer_samples[index, ]$ProteinExpression$`__deferred`$uri
    
    uri <- paste0(cancLink, "?$format=json&$filter=PROTEIN_ID eq ", protein_id)
    
    uriData <- as.data.frame(getAPIData(uri)) |> mutate(
      condition = cancer_samples[index, ]$subtype
    )
    
    temp <- list(uriData)
    cancer_exp_data <- c(cancer_exp_data, temp)
  }
  
  #Combines cancer data and filters useful columns
  cancer_exp <- bind_rows(cancer_exp_data)
  
  cancer_exp <- cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, condition) |> 
    mutate(expression = as.numeric(EXPRESSION))
  
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  plotData$condition <- factor(plotData$condition, levels = c("normal", "ERPR", "HER2", "TN"))
  
  levels(plotData$condition) <- c("normal", "ERPR", "HER2", "TNBC")

  return(plotData)
}
```

Plotting

```{r}
plotProteinSubtype <- function(protein_name, normal_samples, cancer_samples){
  
  #Gets data to plot the ponts
  plotData <- pointsProteinSubtype(protein_name, normal_samples, cancer_samples)
  
  comps <- list(
    c("normal", "ERPR"),
    c("normal", "HER2"),
    c("normal", "TNBC")
  )

  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = condition, y = expression, fill = condition)) +
    geom_boxplot(outliers = F) + 
    labs(
      title = paste(protein_name, "Protein Expression in Breast Cancer Subtypes"),
      x = "condition",
      y = "protein expression"
    ) + 
    geom_signif(
      comparisons = comps,
      map_signif_level  = T,
      test.args = list(paired = F),
      step_increase = 0.08,
    )
}
```

### Subtype

```{r}
pointsProteinListSubtype <- function(protein_list, normal_samples, cancer_samples){
  #Sets up the points to plot
  
  #Gets list of protein ids to use in api call
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  #Iterates through each sample's protein expression data
  normal_exp_data <- list()
  for(normLink in normal_samples$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #Gets API Call
    temp <- list(getAPIData(uri))
    normal_exp_data <- c(normal_exp_data, temp)
  }
  
  #Combines all the expression data together into a single data frame
  normal_exp <- bind_rows(normal_exp_data)
  
  #Filters out only exp data from a specific calculation method and only selects relevant columns
  normal_exp <- normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #Cancer Data
  #Iterates through each sample's protein expression data
  cancer_exp_data <- list()
  
  nCanc <- dim(cancer_samples)[1]
  
  #Iterates through each cancer sample
  for(index in 1:nCanc){
    cancLink <- cancer_samples[index, ]$ProteinExpression$`__deferred`$uri
    
    uri <- paste0(cancLink, "?$format=json&$filter=")
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    uriData <- getAPIData(uri) |> mutate(
      condition = cancer_samples[index, ]$subtype
    )
    
    temp <- list(uriData)
    cancer_exp_data <- c(cancer_exp_data, temp)
  }
  
  #Combines cancer data and filters useful columns
  cancer_exp <- bind_rows(cancer_exp_data)
  
  cancer_exp <- cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID, condition) |> 
    mutate(expression = as.numeric(EXPRESSION))
  
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  plotData$condition <- factor(plotData$condition, levels = c("normal", "ERPR", "HER2", "TN"))
  plotData$PROTEIN_ID <- factor(plotData$PROTEIN_ID)
  
  return(plotData)
}
```

Plotting

```{r}
plotProteinListSubtype <- function(protein_list, normal_samples, cancer_samples, protein_list_name){
  
  protein_id_list <- sapply(protein_list, function(data) proteinNameToID(data)[1])
  
  #Gets data to plot the ponts
  plotData <- pointsProteinListSubtype(protein_list, normal_samples, cancer_samples)
  
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_list_name[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list_name)
  
  #Gets different conditions in proper order
  condLevels <- levels(plotData$condition)
  
  #Makes an interaction that gives every combination between protein and condition to label each box plot
  plotData$interaction <- factor(
    interaction(plotData$protein_name, plotData$condition), 
    levels = c(t(sapply(condLevels, function(d) paste(protein_list_name, d, sep="."))))#Orders the factor levels for boxplot
    )
  
  #Generates labels for the x axis by making a central boxplot have the protein name and keeping the others blank
  xlabels <- c()
  for(protein in protein_list_name){
    xlabels <- c(xlabels, " ", protein, " ", " ")
  }
  
  #Only compare proteins with normal and cancer samples
  valid_list <- c()
  for(index in 1:length(protein_list_name)){
    protein <- protein_list_name[index]
  
    plotData_protein <- plotData |> filter(protein_name == protein)
    levelsPlot <- unique(plotData_protein$condition)
    if(length(levelsPlot) == 4){
      valid_list <- c(valid_list, protein)
    }
  }
  
  nProt <- length(valid_list)
  
  #Creates list of comparisons to make by for each protein, getting the comparisons to be made
  temp <- as.data.frame(sapply(condLevels, function (d) paste(valid_list, d, sep=".")))
  comps <- list()
  for(i in 1:length(valid_list)){
    temp1 <- list(c(temp[i, "normal"], temp[i, "ERPR"]), 
                  c(temp[i, "normal"], temp[i, "HER2"]),
                  c(temp[i, "normal"], temp[i, "TN"])
                  )
    comps <- c(comps, temp1)
  }
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = interaction, y = expression, fill = condition)) +
    scale_x_discrete(
      labels = xlabels 
    ) +
    geom_boxplot(outliers = F) + 
    labs(
    title = "Protein Expression in Breast Cancer Subtypes",
    x = "Protein Name",
    y = "Protein Expression"
  ) + 
    geom_signif(
      data = plotData |> filter(protein_name %in% valid_list),
      comparisons = comps,
      map_signif_level  = T,
      y_position = c(rep(c(10, 10.4, 10.8), nProt)),
    ) + 
    theme_classic()
  
}
```

### Subtype Isoform

```{r}
pointsProteinListIsoSubtype <- function(protein_list, normal_samples, cancer_samples){
  #Sets up the points to plot
  
  #Gets list of protein ids to use in api call
  protein_id_list <- c()
  protein_name_list <- c()
  for(protein in protein_list)
  {
    iso_list <- proteinNameToID(protein)
    protein_id_list <- c(protein_id_list, iso_list)
    protein_name_list <- c(protein_name_list, rep(protein, length(iso_list)))
  }
  
  #count <- 1
  
  #Iterates through each sample's protein expression data
  normal_exp_data <- list()
  for(normLink in normal_samples$ProteinExpression$`__deferred`$uri){
    #Makes the first part of the API URI
    uri <- paste0(normLink, "?$format=json&$filter=")
    #Joins filters for each protein ID
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    #Joins them together
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    #GEta API Call
    temp <- list(getAPIData(uri))
    normal_exp_data <- c(normal_exp_data, temp)
    
    #count <- count +1
    #print(count)
  }
  
  normal_exp <- bind_rows(normal_exp_data)
  
  
  normal_exp <- normal_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID) |> 
    mutate(condition = "normal", expression = as.numeric(EXPRESSION))
  
  #CAncer Data
  #Iterates through each sample's protein expression data
  cancer_exp_data <- list()
  
  nCanc <- dim(cancer_samples)[1]
  
  for(index in 1:nCanc){
    cancLink <- cancer_samples[index, ]$ProteinExpression$`__deferred`$uri
    
    uri <- paste0(cancLink, "?$format=json&$filter=")
    temp <- sapply(protein_id_list, 
                   function(data) paste0("PROTEIN_ID eq ", data)
                   )
    temp <- paste(temp, collapse = " or ")
    uri <- paste0(uri, temp)
    
    uriData <- getAPIData(uri) |> mutate(
      condition = cancer_samples[index, ]$subtype
    )
    
    temp <- list(uriData)
    cancer_exp_data <- c(cancer_exp_data, temp)
    
    #count <- count + 1
    #print(count)
  }
  
  cancer_exp <- bind_rows(cancer_exp_data)
  
  
  
  cancer_exp <- cancer_exp |> 
    filter(CALCULATION_METHOD == calc_method) |> 
    select(EXPRESSION, PROTEIN_ID, condition) |> 
    mutate(expression = as.numeric(EXPRESSION))
  
  
  plotData <- bind_rows(normal_exp, cancer_exp)
  
  plotData$condition <- factor(plotData$condition, levels = c("normal", "ERPR", "HER2", "TN"))
  plotData$PROTEIN_ID <- factor(plotData$PROTEIN_ID)
  
  return(plotData)
}
```

Plotting

```{r}
plotProteinListIsoSubtype <- function(protein_list, normal_samples, cancer_samples){
  
  #Gets data to plot the ponts
  plotData <- pointsProteinListIsoSubtype(protein_list, normal_samples, cancer_samples)
  

  #Gets list of protein ids to use in api call
  protein_id_list <- c()
  protein_name_list <- c()
  for(protein in protein_list)
  {
    iso_list <- proteinNameToID(protein)
    protein_id_list <- c(protein_id_list, iso_list)
    protein_name_list <- c(protein_name_list, rep(protein, length(iso_list)))
  }
  
  plotData$protein_name <- sapply(plotData$PROTEIN_ID, 
                 function(data) protein_name_list[which(data == protein_id_list)])
  
  plotData$protein_name <- factor(plotData$protein_name, levels = protein_list)
  
  n <- plotData |>
    group_by(protein_name, condition) |>
    summarize(n = n()) |>
    pivot_wider(names_from=condition, values_from = n) |>
    mutate(label = paste0(normal,"N ", ERPR,"EP\n", HER2,"H2 ", TN,"TN"))

  nProt <- length(protein_list)
  labels <- n$label
  
  
  #Gets different conditions in proper order
  condLevels <- levels(plotData$condition)
  
  #Makes an interaction that gives every combination between protein and condition to label each box plot
  plotData$interaction <- factor(interaction(plotData$protein_name, plotData$condition), levels = c(t(sapply(condLevels, function(d) paste(protein_list, d, sep=".")))))
  
  #Generates labels for the x axis by making a central boxplot have the protein name and keeping the others blank
  xlabels <- c()
  for(protein in protein_list){
    xlabels <- c(xlabels, " ", protein, " ", " ")
  }
  
    #Creates list of comparisons to make by for each protein, getting the comparisons to be made
  temp <- as.data.frame(sapply(condLevels, function (d) paste(protein_list, d, sep=".")))
  comps <- list()
  for(i in 1:length(protein_list)){
    temp1 <- list(c(temp[i, "normal"], temp[i, "ERPR"]), 
                  c(temp[i, "normal"], temp[i, "HER2"]),
                  c(temp[i, "normal"], temp[i, "TN"])
                  )
    comps <- c(comps, temp1)
  }
  
  #Plots the points in a boxplot
  plotData |> ggplot(aes(x = interaction, y = expression, fill = condition)) +
    scale_x_discrete(
      labels = xlabels #x axis labels
    ) +
    geom_boxplot() + 
    labs(
    title = paste("Protein Expression with Isoforms in Breast Cancer Subtypes"),
    x = "protein name"
  ) + 
    geom_signif( #The significance bars between boxplots
      comparisons = comps,
      map_signif_level  = T,
      y_position = c(rep(c(10, 10.5, 11), nProt)), #Sets the y levels of the signif
      vjust = 0.5
    ) +
    annotate("text", x= 2+4*(0:(nProt-1)), y=5, label=labels) + #num of data points
    coord_cartesian(ylim = c(4, 11.7))
  
}
```

# Control Variables

```{r}
#List of proteins to analyze
protein_list <- c("CD97", "CD109", "CD151", "CD166", "CD276", "EPCAM")
full_protein_list <- c("CD9", "CD97", "CD109", "MUC18","CD151", "CD166", "CD276", "EPCAM")

half_1_protein_list <- c("CD9", "CD97", "CD109", "MUC18")
half_2_protein_list <- c("CD151", "CD166", "CD276", "EPCAM", "FGFR4")

half_1_protein_name <- c("CD9", "CD97", "CD109", "CD146")
half_2_protein_name <- c("CD151", "CD166", "CD276", "EPCAM", "FGFR4")
#Picks calculation method used in protein expression (can give slightly different values)
calc_method = 1

BC_protein_list <- c("CD9", "CD151", "CD276", "EPCAM")
NSCLC_protein_list <- c("CD109", "CD166", "CD276", "EPCAM")

BC_protein_name <- c("CD9", "CD151", "CD276", "EPCAM")
NSCLC_protein_name <- c("CD109", "CD166", "CD276", "EPCAM")
```

# API Call to Proteonomics DB

Maybe it would be easier to go top down

Gets Lung Tissue Data from API calls

```{r}
tissue_data <- getAPIData("https://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Tissue?$format=json")

#Gets all the tissues associated  with the lung and their correesponding links to associated samples
lung_data <- tissue_data |> filter(SAP_SYNONYM == "lung")

#Gets other cancer data from manual searching
cancer_lung_tissues <- tissue_data |> filter(SAP_SYNONYM != "lung" & grepl("lung", NAME) & (grepl("cancer", NAME) | grepl("carcinoma", NAME) ) & TISSUE_ID != "BTO:0000189" & TISSUE_ID != "BTO:0002206")


lung_data <- bind_rows(lung_data, cancer_lung_tissues)

#iterated through each sample to get their data
lung_sample_data <- list()
for(sampLink in lung_data$Sample$`__deferred`$uri){
  temp <- list(getAPIData(paste0(sampLink, "?$format=json")))
  lung_sample_data <- c(lung_sample_data, temp)
}

#Removes empty elements
lung_sample_data <- Filter(length, lung_sample_data)

#Manual Sorting
#Normal Expression
# The second list element is labeled "lung", so it could have normal samples compared to the other elements from cancer cell lines
lung_lung_data <- lung_sample_data[[2]]

#Filters for non-diseased samples and only selects relevant columns
lung_normal_data <- lung_lung_data |> filter(DISEASE == "", TAXCODE == 9606)


#Cancer Data
#Cancer data from tissue data
lung_cancer_data <- lung_lung_data |> filter(DISEASE != "", TAXCODE == 9606)

```

Now to plot the protein expression data in a boxplot

```{r}

# CD97isoforms <- proteinNameToID("CD97")
# 
# plotProteinLung("CD97", lung_normal_data, lung_cancer_data)
# 
# plotProteinLung(CD97isoforms[2], lung_normal_data, lung_cancer_data)
# 
# plotProteinLung(CD97isoforms[3], lung_normal_data, lung_cancer_data)
# 
# plotProteinListLungIso(c("CD97"), lung_normal_data, lung_cancer_data)


```

Now for a different protein

```{r}
# #CD109
# protein_id = 67842
# 
# CD109isoforms <- proteinNameToID("CD109")
# 
# plotProteinLung("CD109", lung_normal_data, lung_cancer_data)
# 
# plotProteinLung(CD109isoforms[2], lung_normal_data, lung_cancer_data)

```

The protein expressions for the proteins of interest are plotted in a boxplot

```{r}
#plotProteinListLung(protein_list, lung_normal_data, lung_cancer_data)

#plotProteinListLungIso(protein_list, lung_normal_data, lung_cancer_data)

#plotProteinListLung(protein_list, lung_normal_data, lung_cancer_data)

#plotProteinListLungIso(tmp_protein_list, lung_normal_data, lung_cancer_data)


plotProteinListLung(half_1_protein_list, lung_normal_data, lung_cancer_data, half_1_protein_name)
plotProteinListLung(half_2_protein_list, lung_normal_data, lung_cancer_data, half_2_protein_name)

#plotProteinListLungIso(half_1_protein_list, lung_normal_data, lung_cancer_data)
#plotProteinListLungIso(half_2_protein_list, lung_normal_data, lung_cancer_data)
```

### Subtype

```{r}
#fje
adeno_data <- tissue_data |> filter(
  (grepl("lung", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM))) & 
    grepl("adeno", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM))) & 
    !grepl("mouse", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM)))) |
    NAME %in% c("A-549 cell", "NCI-H322M cell")
)

squamous_data <- tissue_data |> filter(
  (grepl("lung", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM))) & 
    grepl("squamous", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM))) & 
    !grepl("mouse", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM)))) |
    NAME %in% c("NCI-H226 cell")
)

LLC_data <- tissue_data |> filter(
  (grepl("large cell", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM))) & 
    grepl("lung", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM))) & 
    !grepl("mouse", str_to_lower(paste(NAME, DEFINITION, SYNONYM, SAP_SYNONYM)))) |
    NAME %in% c("NCI-H460 cell")
)


lung_cancer_list <- list(adeno_data, squamous_data, LLC_data)
lung_subtype_list <- c("adeno", "squamous", "large cell")

  #DEBUGGING
  sample_list <- list()
  for(subtype in lung_cancer_list)
  {
    subtype_sample <- data.frame()
    for(link in subtype$Sample$`__deferred`$uri){
      temp <- as.data.frame(getAPIData(paste0(link, "?$format=json")))
      subtype_sample <- bind_rows(subtype_sample, temp)
    }
    sample_list <- c(sample_list, list(subtype_sample))
  }
  
  #Samples show that samples from a specific project (Procancer 949 cell lines) have consistently low protein expressions across all proteins, so will be excluded
  for(i in 1:length(sample_list)){
    sample_list[[i]] <- sample_list[[i]] |> 
      filter(TREATMENT == "" | grepl("none", TREATMENT) | TREATMENT == "NA") |>
      filter( !(SAMPLE_ID >= 1032827 & SAMPLE_ID <= 1038412))
}

 #samples <- bind_rows(sample_list)
# samples <- samples |> filter(TREATMENT == "" | grepl("none", TREATMENT) | TREATMENT == "NA")
# 
# exclude_samples <- samples |> filter(SAMPLE_ID >= 1032827 & SAMPLE_ID <= 1038412)
# exclude_samples <- exclude_samples$SAMPLE_ID

```

Plotting Group

```{r}
plotProteinListLungSubtype(half_1_protein_list, lung_normal_data, sample_list, lung_subtype_list, half_1_protein_name)

plotProteinListLungSubtype(half_2_protein_list, lung_normal_data, sample_list, lung_subtype_list, half_2_protein_name)
```

Plotting individual

```{r}
plotProteinLungSubtype("CD109", lung_normal_data, sample_list, lung_subtype_list)
ggsave("plots/CD109_NSCLC_protein_expression.png")

plotProteinLungSubtype("CD151", lung_normal_data, sample_list, lung_subtype_list)
ggsave("plots/CD151_NSCLC_protein_expression.png")

plotProteinLungSubtype("CD166", lung_normal_data, sample_list, lung_subtype_list)
ggsave("plots/CD166_NSCLC_protein_expression.png")

plotProteinLungSubtype("EPCAM", lung_normal_data, sample_list, lung_subtype_list)
ggsave("plots/EPCAM_NSCLC_protein_expression.png")

plotProteinLungSubtype("CD276", lung_normal_data, sample_list, lung_subtype_list)
ggsave("plots/CD276_NSCLC_protein_expression.png")


```

Removing batch effects?? Not much difference has been found in removing batch effects between different quantification/acquisition mode (MS/MS Detector)

```{r}
# temp <- pointsProteinLungSubtype("CD109", lung_normal_data, sample_list, lung_subtype_list)
# exp_sample_lung <- temp |> left_join(lung_samp_2, join_by(SAMPLE_ID))
# lung_sample <- bind_rows(sample_list)
# lung_samp_2 <- bind_rows(lung_sample, lung_normal_data)
# exp_sample_lung <- temp |> left_join(lung_samp_2, join_by(SAMPLE_ID))
# exp_sample_lung <- exp_sample_lung |> select(!`__metadata`)
# exp_sample_lung <- exp_sample_lung |> select(!EXPRESSION)
# exp_samp_lung <- exp_sample_lung |> select(SAMPLE_ID:DESCRIPTION, QUANTIFICATION_METHOD:ACQUISITION_MODE)
# design <- exp_samp_lung$condition
# batch <- exp_samp_lung$MS_DETECTOR
# batch2 <- exp_samp_lung$QUANTIFICATION_METHOD
# x <- exp_samp_lung$expression
# 
# x <- t(x)
# 
# 
# levels(design) <- 1:4
# design <- as.matrix(as.numeric(design))
# x_noBatch <- removeBatchEffect(x, batch = batch, batch2 = batch2, design = design)
# 
# toPlot <- as.data.frame(t(x_noBatch) |> cbind(design))
# colnames(toPlot) <- c("expression", "condition") 
# 
# toPlot$condition <- factor(toPlot$condition)
# levels(toPlot$condition) <- c("normal", "adeno", "SqCC", "LCC")
# 
# toPlot |> ggplot(aes(x = condition, y = expression, fill = condition)) + geom_boxplot()
```

# Breast Cancer Data

Analyzes the breast cancer data by filtering the breast tissue data from the total tissue data and using manual searching to find normal or cancer data

```{r}
#View breast samples

breast_tissue <- tissue_data |> 
  filter(grepl("breast", NAME) | grepl("breast", SAP_SYNONYM)) |> 
  filter(!(NAME %in% c("MCF-10A cell", "HMT-3522 cell", "breast hyperplasia cell", "HMELonza"))) #Removing problematic tissues

breast_sample_data <- list()
for(sampLink in breast_tissue$Sample$`__deferred`$uri){
  temp <- list(getAPIData(paste0(sampLink, "?$format=json")))
  
  breast_sample_data <- c(breast_sample_data, temp)
}

#Removes empty elements
breast_tissue <- breast_tissue[!sapply(breast_sample_data, is_empty), ]

breast_sample_data <- Filter(length, breast_sample_data)

#Removes any samples with treatments
breast_sample_data <- lapply(breast_sample_data, 
                             function(data) filter(data, 
(TREATMENT == "" | TREATMENT == "NA" | grepl("none", TREATMENT))
                                                   )
                             )
```

```{r}



#Obtaining normal breast data from potential tissues (some are empty)
HMEpC <- breast_sample_data[[11]]
breastData <- breast_sample_data[[2]]
MCF10F <- getAPIData("http://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Tissue('BTO%3A0003900')/Sample?$format=json")
HUMA7 <- getAPIData("http://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Tissue('BTO%3A0004127')/Sample?$format=json")
HUMAN7 <- getAPIData("http://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Tissue('BTO%3A0004865')/Sample?$format=json")  


#Gets the data frame of sample data for normal and cancer
breastNorm <- breastData |> filter(DISEASE_ID == "")

breast_normal_data <- bind_rows(HMEpC, breastNorm, MCF10F, HUMA7, HUMAN7)

breastCancer <- breastData |> filter(DISEASE_ID != "")

breast_cancer_data <- breastCancer

for(index in 1:length(breast_sample_data)){
  breast_sample_data[[index]]$ELECTROPHORESIS <- as.character(breast_sample_data[[index]]$ELECTROPHORESIS)
}

breast_cancer_data <- bind_rows(breast_cancer_data, breast_sample_data[c(-2, -11)])

#plotDataB <- pointsProteinList(protein_list, breast_normal_data, breast_cancer_data)

#plotProteinList(protein_list, breast_normal_data, breast_cancer_data)

#plotProteinListIso(protein_list, breast_normal_data, breast_cancer_data)

plotProteinList(half_1_protein_list, breast_normal_data, breast_cancer_data, half_1_protein_name)
plotProteinList(half_2_protein_list, breast_normal_data, breast_cancer_data, half_1_protein_name)

#plotProteinListIso(half_1_protein_list, breast_normal_data, breast_cancer_data)
#plotProteinListIso(half_2_protein_list, breast_normal_data, breast_cancer_data)
```

## Subtype Analysis

Breast cancer data is in the project "Tyanova_ncomms_2016", with the subtype information available in the paper. Normal data can be taken from the previous data

```{r}

#Uses the project id of the project to get the experiment data (id = 4473)
project_data_BC <- getAPIData("https://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Project?$format=json&$filter=PUBMEDID eq '26725330'")

#Gets experiment data (experiment id = 13219)
experiment_data_BC <- getAPIData("https://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Project(4473)/Experiment?$format=json")

#Gets sample data  for the project
sample_data_BC <- getAPIData("https://www.proteomicsdb.org/proteomicsdb/logic/api_v2/api.xsodata/Experiment(13219)/Sample?$format=json")

#Gets subtype data for the project samples
subtype_BC_data <- read_delim("data/BC_Proteomics_Subtype.txt", delim = " ")

sample_subtype_BC <- sample_data_BC |> select(NAME, ProteinExpression)

#Gets numeric value of sample
sample_subtype_BC$sample_num <- sapply(sample_subtype_BC$NAME, function(data) substr(data, start = 3, stop = 4)) |> str_replace("_", "") |> as.numeric()

#Gets numeric value of sample to compare to subtype
subtype_BC_data <- subtype_BC_data[-27, ]
subtype_BC_data$`Sample_#` <- as.numeric(subtype_BC_data$`Sample_#`)

#Adds the subtype for each sample
sample_subtype_BC$subtype <- sapply(sample_subtype_BC$sample_num, 
                 function(data) subtype_BC_data$Clinical_classification[which(data == subtype_BC_data$`Sample_#`)])



plotProteinListSubtype(half_1_protein_list, breast_normal_data, sample_subtype_BC, half_1_protein_name)
plotProteinListSubtype(half_2_protein_list, breast_normal_data, sample_subtype_BC, half_2_protein_name)

#plotProteinListIsoSubtype(half_1_protein_list, breast_normal_data, sample_subtype_BC)
#plotProteinListIsoSubtype(half_2_protein_list, breast_normal_data, sample_subtype_BC)
```

```{r}
#| warning: false

plotProteinSubtype("CD9", breast_normal_data, sample_subtype_BC)
ggsave("plots/CD9_BC_protein_expression.png")

plotProteinSubtype("CD151", breast_normal_data, sample_subtype_BC)
ggsave("plots/CD151_BC_protein_expression.png")

plotProteinSubtype("CD166", breast_normal_data, sample_subtype_BC)
ggsave("plots/CD166_BC_protein_expression.png")

plotProteinSubtype("EPCAM", breast_normal_data, sample_subtype_BC)
ggsave("plots/EPCAM_BC_protein_expression.png")

plotProteinSubtype("CD276", breast_normal_data, sample_subtype_BC)
ggsave("plots/CD276_BC_protein_expression.png")


```

```{r}
plotProteinListLung(NSCLC_protein_list, lung_normal_data, lung_cancer_data, NSCLC_protein_name)

plotProteinListLungSubtype(NSCLC_protein_list, lung_normal_data, sample_list, lung_subtype_list, NSCLC_protein_name)

plotProteinList(BC_protein_list, breast_normal_data, breast_cancer_data, BC_protein_name)

plotProteinListSubtype(BC_protein_list, breast_normal_data, sample_subtype_BC, BC_protein_name)
```
